---
title: Prepare Data Layers
subtitle: Lorem Ipsum
description: Convert the needed data to parquet format by either downloading them from online or from loca data.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",    # https://github.com/tidyverse/tidyverse/
    "sf",           # https://github.com/r-spatial/sf/
    "arcgislayers", # https://github.com/R-ArcGIS/arcgislayers/
    "arrow",        # https://github.com/apache/arrow/tree/main/r
    "geoarrow",     # https://github.com/geoarrow/geoarrow-r/
    "janitor",      # https://github.com/sfirke/janitor/
    "mapgl",        # https://github.com/walkerke/mapgl/
    "here",         # https://github.com/r-lib/here/
    "fs"            # https://github.com/r-lib/fs/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(sf)
library(arcgislayers)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
library(fs)
```

## Project Configurations

### Arrow Initialization

```r
# Only run if this is the first time using `arrow` package on this device
arrow::install_arrow()
```

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source     = here::here("_src"), # Source custom functions
  raw        = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed  = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered   = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results    = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base      = here::here(dirs$processed, "base_layers"),
  parcels   = here::here(dirs$processed, "parcels_taz"),
  landuse   = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access    = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```


## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "as_geoparquet_table.R")) # Write Geoparquet metadata
source(here::here(dirs$source, "download_feature_layer.R")) # Download ArcGIS Feature Layers
source(here::here(dirs$source, "process_local_layer.R")) # Load local SHP, GDB, GeoJSON Layers
source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

# Base Layers (The Canvas)

## Administrative Boundaries

### Municipal Boundaries

```{r}
#| label: muni-bounds
fp_municipal_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/UtahMunicipalBoundaries/FeatureServer/0",
  name = "UtahMunicipalBoundaries",
  save_dir = sub_dirs$base, # Saves to _data/processed/base_layers/
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Municipal Boundaries
preview_layer(fp_municipal_bound)
```

### County Boundaries

```{r}
#| label: county-bounds
fp_county_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/UtahCountyBoundaries/FeatureServer/0",
  name = "UtahCountyBoundaries",
  save_dir = sub_dirs$base,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check County Boundaries
preview_layer(fp_county_bound)
```

### Regional Boundaries (WFRC/MAG)

```{r}
#| label: region-bounds
fp_regional_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/RegionalBoundaryComponents/FeatureServer/0",
  name = "RegionalBoundaryComponents",
  # query = "PlanOrg IN ('MAG MPO', 'WFRC MPO')",
  save_dir = sub_dirs$base,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Regional Boundaries
preview_layer(fp_regional_bound)
```

## The "Target" Universe: Parcels

### REMM Parcels

```{r}
# REMM Parcels (from GDB)
fp_remm_parcels <- process_local_layer(
  path = "remm_base_year_20221003/remm_base_year.gdb.zip",
  layer = "parcels",
  name = "REMM_BY_parcels",
  save_dir = sub_dirs$parcels,
  # partitioning = "co_name",
  compression = "zstd",
  crs  = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check REMM Parcels (Limit to 1000 so browser doesn't crash)
preview_layer(fp_remm_parcels, limit = 1000)
```

### REMM Buildings

```{r}
# REMM Buildings (from GDB)
fp_remm_buildings <- process_local_layer(
  path = "remm_base_year_20221003/remm_base_year.gdb.zip",
  layer = "buildings",
  name = "REMM_BY_buildings",
  save_dir = sub_dirs$parcels,
  # partitioning = "building_type",
  compression = "zstd",
  crs  = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check REMM Buildings (Limit to 1000 so browser doesn't crash)
preview_layer(fp_remm_buildings, limit = 1000)
```

### Wasatch Front - Traffic Analysis Zones (TAZ)

```{r}
#| label: taz-zones
fp_wf_taz <- process_local_layer(
  path = "WFv920_TAZ.zip",
  name = "WFv920_TAZ",
  save_dir = sub_dirs$parcels,
  crs  = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check WF-TDM TAZs
preview_layer(fp_wf_taz)
```

### Utah Statewide - Traffic Analysis Zones (TAZ)

```{r}
#| label: taz-zones
fp_ustm_taz <- process_local_layer(
  path = "Statewide_TAZ.zip",
  name = "Statewide_TAZ",
  save_dir = sub_dirs$parcels,
  crs  = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check WF-TDM TAZs
preview_layer(fp_ustm_taz)
```

## Land Use & Zones

### Wasatch Choice Vision Centers

```{r}
#| label: wcv-centers
fp_wcv_centers <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/WCV_Centers_and_Regional_Land_Uses/FeatureServer/0",
  name = "WCV_Centers_and_Regional_Land_Uses",
  save_dir = sub_dirs$landuse,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Wasatch Choice Vision Centers
preview_layer(fp_wcv_centers, column = "center_type")
```

### Utah Qualified Opportunity Zones (QOZ).

```{r}
#| label: qoz
fp_utah_qoz <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/Utah_Qualified_Opportunity_Zones/FeatureServer/0",
  name = "Utah_Qualified_Opportunity_Zones",
  save_dir = sub_dirs$landuse,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Utah Qualified Opportunity Zones
preview_layer(fp_utah_qoz)
```

# Transportation Network (Data & Buffers)

## Transit Network

### Transit Routes

```{r}
#| label: transit-routes

fp_transit_routes <- download_feature_layer(
  url = "https://maps.rideuta.com/server/rest/services/Hosted/UTA_Routes_and_Most_Recent_Ridership/FeatureServer/0",
  name = "UTA_Routes_and_Most_Recent_Ridership",
  save_dir = sub_dirs$transport,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Transit Routes
preview_layer(fp_transit_routes, column = "routetype")
```

### Transit Stops

```{r}
#| label: transit-stops
fp_transit_stops <- download_feature_layer(
  url = "https://maps.rideuta.com/server/rest/services/Hosted/UTA_Stops_and_Most_Recent_Ridership/FeatureServer/0",
  name = "UTA_Stops_and_Most_Recent_Ridership",
  save_dir = sub_dirs$transport,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Transit Stops
preview_layer(fp_transit_stops)
```

## Active Transportation

```{r}
#| label: bikeways

fp_bikeways <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Bikeways/FeatureServer/0",
  name = "Bikeways",
  save_dir = sub_dirs$transport,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Utah Bikeways
preview_layer(fp_bikeways, column = "facility1", limit = 1000)
```

## Roadway Network

## Freeway Exits

```{r}
#| label: freeway-exits

fp_freeway_exits <- download_feature_layer(
  url = "https://services.arcgis.com/pA2nEVnB6tquxgOW/arcgis/rest/services/Freeway_Exit_Locations/FeatureServer/0",
  name = "Freeway_Exit_Locations",
  save_dir = sub_dirs$transport,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Freeway Exits
preview_layer(fp_freeway_exits)
```

# Access to Opportunity (ATO)

```{r}
#| label: taz_ato
#| eval: false
#| code-fold: true

fp_taz_ato <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/AccessToOpportunities_gdb/FeatureServer/0",
  name = "TAZ_AccessToOpportunities_OY2023",
  save_dir = sub_dirs$access,
  crs = CRS_WGS
)
```

```{r}
#| label: taz_ato

# ATO Source: "\\ModelAce\ModelAce-E\1 - TDM\1 - Official Release (full run)\v9x\v9.2\v9.2.0\Scenarios\OY_2023\5_AssignHwy\4_Summaries\WFv920_OY_2023_Access_to_Opportunity.csv"

# 1. Define Paths
fp_taz_ato <- here::here(sub_dirs$access, "TAZ_AccessToOpportunities_OY2023.parquet")
fp_ato_csv <- here::here(dirs$raw, "WFv920_OY_2023_Access_to_Opportunity.csv")

# 2. Conditional Processing
if (file.exists(fp_taz_ato)) {
  message("✅ TAZ ATO layer already exists: ", fp_taz_ato)
} else {

  # Single Pipeline: Read TAZ -> Clean -> Join(Read CSV -> Clean) -> Save
  arrow::read_parquet(fp_wf_taz) |>           # Read LHS (Geometry) to memory
    sf::st_as_sf(crs = CRS_WGS) |>           # Convert to SF (EPSG:4326)
    janitor::clean_names() |>                 # Clean LHS names
    dplyr::left_join(
      # Nested Pipeline: Read RHS (Attributes) -> Clean Names
      arrow::read_csv_arrow(fp_ato_csv) |>
        janitor::clean_names(),
      by = "co_tazid"
    ) |>
    as_geoparquet_table() |>                  # Convert back to Arrow with Metadata
    arrow::write_parquet(fp_taz_ato)

  message("✅ Joined and saved to: ", fp_taz_ato)
}
```

```{r}
#| label: verify
#| eval: false

# Check Access to Opportunities
preview_layer(fp_taz_ato, column = "job_by_auto")
```

# Community Amenities (Data & Buffers)

## Education

### K-12 Schools (Public/Charter)

```{r}
#| label: schools-k12
fp_edu_k12 <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Schools_PreKto12/FeatureServer/0",
  name = "Schools_PreKto12",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check K-12 Schools
preview_layer(fp_edu_k12)
```

### Higher Education

```{r}
#| label: schools-higher
fp_edu_higher <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Schools_HigherEducation/FeatureServer/0",
  name = "Schools_HigherEducation",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Higher Education
preview_layer(fp_edu_higher)
```

## Public Health & Safety

### Healthcare Facilities, Hospitals.

```{r}
#| label: health-facilities
fp_health_fac <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/LicensedHealthCareFacilities/FeatureServer/0",
  name = "LicensedHealthCareFacilities",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Healthcare Facilities
preview_layer(fp_health_fac)
```

### Childcare Centers

```{r}
#| label: childcare-facilities
fp_child_care <- process_local_layer(
  path = "HIFLD OPEN Child Care Centers/child-care-centers-geojson.zip",
  name = "ChildCareFacilities",
  query = "SELECT * FROM Child_Care_Centers WHERE STATE = 'UT'",
  read_mode = "zip",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Childcare Facilities
preview_layer(fp_child_care)
```

## Lifestyle & Services

### Groceries Stores

```{r}
#| label: grocery-store
fp_grocery_store <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/UtahGroceryAndFoodStores_DAF/FeatureServer/0",
  name = "UtahGroceryAndFoodStores_DAF",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Grocery Stores
preview_layer(fp_grocery_store)
```

### Community Centers

```{r}
#| label: community-center
fp_community_center <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/Residential_Accessibility_WFL1/FeatureServer/412", # TODO: use UGRC
  name = "Community_Centers",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Community Centers
preview_layer(fp_community_center)
```

### Parks and Open Spaces (Walk Access)

```{r}
#| label: park-access
fp_park_access <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/ParkEdits/FeatureServer/0",
  name = "WalkAccesstoParks_10Min",
  save_dir = sub_dirs$amenities,
  crs = CRS_WGS
)
```

```{r}
#| label: verify
#| eval: false

# Check Walk Access to Parks and Open Spaces
preview_layer(fp_park_access)
```

# Quality Control & Manifest

## File Verification

* List all files in `_data/raw` and `_data/processed`.
* Quick sanity check on file sizes (ensure non-zero bytes).

```{r}
#| label: verify
#| eval: false

# List processed files to ensure structure is correct
list.files(dirs$processed, recursive = TRUE)
```
