---
title: Prepare Data Layers
subtitle: Lorem Ipsum
description: Convert the needed data to parquet format by either downloading them from online or from loca data.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arcgis",      # https://github.com/R-ArcGIS/arcgis/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here"         # https://github.com/r-lib/here/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(sf)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arcgislayers)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Database Setup/Initialization

```r
# Only run if this is the first time using `arrow` package on this device
arrow::install_arrow()
```

```{r}
#| label: setup-db-init

# Initialize In-memory database
conn <- DBI::dbConnect(duckdb::duckdb())

# install and load spatial extension
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
```

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  raw = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed = here::here("_data/processed"), # Intermediate Output: Standardized Parquets & Buffers
  results = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~dir.create(., recursive = TRUE, showWarnings = FALSE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base   = file.path(dirs$processed, "base_layers"),
  parcels = file.path(dirs$processed, "parcels_taz"),
  landuse = file.path(dirs$processed, "landuse_zones"),
  transport = file.path(dirs$processed, "transportation"),
  amenities = file.path(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~dir.create(., recursive = TRUE, showWarnings = FALSE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_PROJ <- "EPSG:3566" # NAD83 / Utah Central (Projected - ftUS)
CRS_VIEW <- "EPSG:3857" # WGS 84 / Pseudo-Mercator (Projected - Meters)
CRS_WGS  <- "EPSG:4326" # WGS84 (Geographic - Lat/Lon)
```


## Helper Functions

```{r}
#| label: load-helpers

source(here::here("_src/download_feature_layer.R")) # Download ArcGIS Feature Layers
source(here::here("_src/process_local_layer.R")) # Load local SHP, GDB, GeoJSON Layers
source(here::here("_src/preview_layer.R")) # Preview (geo)parquet files using mapgl
```

# Base Layers (The Canvas)

## Administrative Boundaries

### Municipal Boundaries

```{r}
#| label: muni-bounds
fp_municipal_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/UtahMunicipalBoundaries/FeatureServer/0",
  name = "UtahMunicipalBoundaries",
  save_dir = sub_dirs$base, # Saves to _data/processed/base_layers/
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Municipal Boundaries
preview_layer(fp_municipal_bound, color = "purple")
```

### County Boundaries

```{r}
#| label: county-bounds
fp_county_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/UtahCountyBoundaries/FeatureServer/0",
  name = "UtahCountyBoundaries",
  save_dir = sub_dirs$base,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check County Boundaries
preview_layer(fp_county_bound, color = "purple")
```

### Regional Boundaries (WFRC/MAG)

```{r}
#| label: region-bounds
fp_regional_bound <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/RegionalBoundaryComponents/FeatureServer/0",
  name = "RegionalBoundaryComponents",
  # query = "PlanOrg IN ('MAG MPO', 'WFRC MPO')",
  save_dir = sub_dirs$base,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Regional Boundaries
preview_layer(fp_regional_bound, color = "purple")
```

## The "Target" Universe: Parcels

* Note: This is likely your largest file.
* Download/Load Parcels.
* Standardize Column Names (clean for SQL).
* Save parcels.parquet.

### Land Parcels

```{r}
# Parcels (from GDB)
fp_parcels <- process_local_layer(
  relative_path = "remm_base_year_20221003/remm_base_year.gdb",
  output_name   = "parcels",
  layer_name    = "Buildings" # Verify this layer name in your GDB
)
```

```{r}
#| label: verify
#| eval: false

# Check Parcels (Limit to 1000 so browser doesn't crash)
preview_layer(fp_parcels, color = "orange", limit = 1000)
```

### Traffic Analysis Zones (TAZ).

```{r}
#| label: taz-zones
fp_wf_taz <- process_local_layer(
  path = "WFv920_TAZ.zip",
  name = "WFv920_TAZ",
  save_dir = sub_dirs$parcels,
  crs  = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check WF-TDM TAZs
preview_layer(fp_wf_taz, color = "orange", limit = 999)
```

## Land Use & Zones

### Wasatch Choice Vision Centers

```{r}
#| label: wcv-centers
fp_wcv_centers <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/WCV_Centers_and_Regional_Land_Uses/FeatureServer/0",
  name = "WCV_Centers_and_Regional_Land_Uses",
  save_dir = sub_dirs$landuse,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Wasatch Choice Vision Centers
preview_layer(fp_wcv_centers, color = "purple")
```

### Utah Qualified Opportunity Zones (QOZ).

```{r}
#| label: qoz
fp_utah_qoz <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/Utah_Qualified_Opportunity_Zones/FeatureServer/0",
  name = "Utah_Qualified_Opportunity_Zones",
  save_dir = sub_dirs$landuse,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Utah Qualified Opportunity Zones
preview_layer(fp_utah_qoz, color = "purple")
```

# Transportation Network (Data & Buffers)

## Transit Network

### Transit Routes

```{r}
#| label: transit-routes
fp_transit_routes <- download_feature_layer(
  url = "https://maps.rideuta.com/server/rest/services/Hosted/UTA_Routes_and_Most_Recent_Ridership/FeatureServer/0",
  name = "UTA_Routes_and_Most_Recent_Ridership",
  save_dir = sub_dirs$transport,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Transit Routes
preview_layer(fp_transit_routes, color = "red") # TODO: missing crs
```

### Transit Stops

```{r}
#| label: transit-stops
fp_transit_stops <- download_feature_layer(
  url = "https://maps.rideuta.com/server/rest/services/Hosted/UTA_Stops_and_Most_Recent_Ridership/FeatureServer/0",
  name = "UTA_Stops_and_Most_Recent_Ridership",
  save_dir = sub_dirs$transport,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Transit Stops
preview_layer(fp_transit_stops, color = "red")
```

## Active Transportation

```{r}
#| label: bikeways
fp_bikeways <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Bikeways/FeatureServer/0",
  name = "Bikeways",
  save_dir = sub_dirs$transport,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Utah Bikeways
preview_layer(fp_bikeways, color = "red") # TODO: mising CRS
```

## Roadway Network

* Download: Highway Interchanges.
* Processing: Generate Interchange buffers.

# Community Amenities (Data & Buffers)

## Education

### K-12 Schools (Public/Charter)

```{r}
#| label: schools-k12
fp_edu_k12 <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Schools_PreKto12/FeatureServer/0",
  name = "Schools_PreKto12",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check K-12 Schools
preview_layer(fp_edu_k12, color = "red")
```

### Higher Education

```{r}
#| label: schools-higher
fp_edu_higher <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/Schools_HigherEducation/FeatureServer/0",
  name = "Schools_HigherEducation",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Higher Education
preview_layer(fp_edu_higher, color = "red")
```

## Public Health & Safety

### Healthcare Facilities, Hospitals.

```{r}
#| label: health-facilities
fp_health_fac <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/LicensedHealthCareFacilities/FeatureServer/0",
  name = "LicensedHealthCareFacilities",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Healthcare Facilities
preview_layer(fp_health_fac, color = "red")
```

### Childcare Centers

```{r}
#| label: health-facilities
fp_child_care <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/Residential_Accessibility_WFL1/FeatureServer/5", # TODO: use UGRC
  name = "ChildCareFacilities",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Childcare Facilities
preview_layer(fp_child_care, color = "red")
```

## Lifestyle & Services

### Groceries Stores

```{r}
#| label: grocery-store
fp_grocery_store <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/Residential_Accessibility_WFL1/FeatureServer/374", # TODO: use UGRC
  name = "Grocery_Stores",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Grocery Stores
preview_layer(fp_grocery_store, color = "red")
```

### Community Centers

```{r}
#| label: community-center
fp_community_center <- download_feature_layer(
  url = "https://services1.arcgis.com/taguadKoI1XFwivx/ArcGIS/rest/services/Residential_Accessibility_WFL1/FeatureServer/412", # TODO: use UGRC
  name = "Community_Centers",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Community Centers
preview_layer(fp_community_center, color = "red")
```

### Parks and Open Spaces

```{r}
#| label: open-spaces
fp_open_spaces <- download_feature_layer(
  url = "https://services1.arcgis.com/99lidPhWCzftIe9K/ArcGIS/rest/services/UtahParksLocal/FeatureServer/0",
  name = "UtahParksLocal",
  save_dir = sub_dirs$amenities,
  crs = CRS_PROJ
)
```

```{r}
#| label: verify
#| eval: false

# Check Parks and Open Spaces
preview_layer(fp_open_spaces, color = "red")
```

# Quality Control & Manifest

## File Verification

* List all files in `_data/raw` and `_data/processed`.
* Quick sanity check on file sizes (ensure non-zero bytes).

```{r}
#| label: verify
#| eval: false

# List processed files to ensure structure is correct
list.files(dirs$processed, recursive = TRUE)
```

## Visual Sample

* Use `mapgl` or `tmap` to visualize one buffer layer over the parcel centroids to ensure projections align.


```{r}
#| label: dis-conn
DBI::dbDisconnect(conn)
```
