---
title: Process Data Layers
subtitle: Lorem Ipsum
description: Process the data to prepare for spatial operations (buffer, isochrones, etc.)
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
execute:
  eval: false
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here",        # https://github.com/r-lib/here/
    "fs",          # https://github.com/r-lib/fs/
    "zip"          # https://github.com/r-lib/zip
    ),
    dependencies = TRUE
  )
```

OR,

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak("renv", dependencies = TRUE) # https://github.com/rstudio/renv/

renv::restore()
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Setup OpenRouteService API

```{r}
#| label: setup-ors

# --- Configuration ---
# TRUE  = Use local Docker instance (Free, Unlimited, Fast)
# FALSE = Use Public Cloud API (Limited tokens, Requires Key)
USE_LOCAL_ORS <- TRUE

if (USE_LOCAL_ORS) {
  # 1. Local Setup
  local_url <- "http://localhost:8080/ors"

  # Set the option that the package looks for
  options(openrouteservice.url = local_url)

  # Local instances ignore the key, but the package sometimes checks for existence.
  # We set a dummy key to prevent errors.
  suppressMessages(ors_api_key("local-instance"))

  message("✅ Using Local ORS instance at: ", local_url)
} else {
  # 2. Remote Setup
  # Important: Clear the URL option so it defaults back to "api.openrouteservice.org"
  options(openrouteservice.url = NULL)

  # Load Key
  readRenviron(here::here("_environment"))

  if (Sys.getenv("ORS_API_KEY") == "") {
    stop("❌ ORS_API_KEY not found in _environment file.")
  } else {
    ors_api_key(Sys.getenv("ORS_API_KEY"))
    message("✅ Using Public ORS API (Key Loaded)")
  }
}
```

```r
# Optionally set ORS Paths
options(
  openrouteservice.paths = list(
      directions = "v2/directions",
      isochrones = "v2/isochrones",
      matrix = "v2/matrix",
      geocode = "geocode",
      pois = "pois",
      elevation = "elevation",
      optimization = "optimization",
      snap = "v2/snap",
      export = "v2/export",
      health = "v2/health",
      status = "v2/status"
  )
)
```

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source = here::here("_src"), # Source custom functions
  raw = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~ fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base = here::here(dirs$processed, "base_layers"),
  parcels = here::here(dirs$processed, "parcels_taz"),
  landuse = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~ fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-iso-subdirs

# Define Sub-directories for Isochrones
iso_dirs <- list(
  walking = here::here(dirs$isochrones, "foot-walking"),
  biking = here::here(dirs$isochrones, "cycling-regular"),
  driving = here::here(dirs$isochrones, "driving-car")
)

purrr::walk(iso_dirs, ~ fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566" # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326" # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "as_geoparquet_table.R")) # Write Geoparquet metadata
source(here::here(dirs$source, "process_isochrones.R")) # Generate isochrones using OpenRouteService API
source(here::here(dirs$source, "create_buffer.R")) # Generate equidistant buffers
source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

# Preprocessing

## Set Regional Boundary

```r
#| label: prep-boundary

# -----------------------------------------------------------------------------
# DOESN'T WORK CURRENTLY BECAUSE {duckspatial} IS STILL UNDER DEVELOPMENT
# -----------------------------------------------------------------------------

fp_regional_bound <- here::here(sub_dirs$base, "RegionalBoundaryComponents.parquet")
fp_boundary_wftdm <- here::here(sub_dirs$base, "Analysis_Boundary_WFRC_MAG.parquet")

if (file.exists(fp_boundary_wftdm)) {
  message("✅ Analysis boundary already exists: ", fp_boundary_wftdm)
} else {

  # 1. Connection Management (duckspatial)
  # Uses ddbs_create_conn for standardized in-memory DB setup
  conn <- duckspatial::ddbs_create_conn()
  on.exit(duckspatial::ddbs_stop_conn(conn), add = TRUE)

  # 2. Extension Management (duckspatial)
  duckspatial::ddbs_load(conn)
  DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")

  # 3. Define Lazy Source (duckplyr)
  # duckplyr works seamlessly with DuckDB connections
  tbl_boundary <- dplyr::tbl(conn, paste0("read_parquet('", fp_regional_bound, "')"))

  # 4. Build Pipeline
  processed_boundary <- tbl_boundary |>

    # Filter using duckplyr (optimized translation)
    duckplyr::filter(plan_org %in% c('MAG MPO', 'WFRC MPO')) |>

    # A. Initial Dissolve
    # MISSING FUNCTION: ddbs_union_agg()
    duckplyr::summarise(geom = ddbs_union_agg(geometry)) |>

    # B. Explode MultiPolygons
    # MISSING FUNCTION: ddbs_dump() returns a list/struct
    duckplyr::mutate(parts = ddbs_dump(geom)) |>

    # Unnest is a tidyr verb, but handled by dbplyr/duckplyr backend
    tidyr::unnest_longer(parts) |>

    # C. Remove Holes (ExteriorRing -> MakePolygon)
    # MISSING FUNCTIONS: ddbs_exterior_ring(), ddbs_make_polygon()
    duckplyr::mutate(
      # Accessing struct field '.geom' from the dump result
      outer_ring = ddbs_exterior_ring(parts$geom),
      clean_poly = ddbs_make_polygon(outer_ring)
    ) |>

    # D. Re-Aggregate and Convert to WKB
    # MISSING FUNCTIONS: ddbs_union_agg(), ddbs_as_wkb()
    duckplyr::summarise(geometry = ddbs_as_wkb(ddbs_union_agg(clean_poly)))

  # 5. Execute & Write (Zero-Copy)
  # We convert the duckplyr lazy object to SQL and execute COPY
  query_sql <- dbplyr::sql_render(processed_boundary)
  DBI::dbExecute(conn, glue::glue("COPY ({query_sql}) TO '{fp_boundary_wftdm}' (FORMAT PARQUET)"))

  message("✅ Analysis boundary created (Duck-Native): ", fp_boundary_wftdm)
}
```

```{r}
#| label: prep-boundary

# Define inputs/outputs
fp_regional_bound <- here::here(
  sub_dirs$base,
  "RegionalBoundaryComponents.parquet"
)
fp_boundary_wftdm <- here::here(
  sub_dirs$base,
  "Analysis_Boundary_WFRC_MAG.parquet"
)

# Define the CRS of your INPUT file 'RegionalBoundaryComponents.parquet'
# Common Utah CRSs: "EPSG:26912" (UTM 12N) or "EPSG:3566" (Utah Central State Plane)
SOURCE_CRS <- CRS_WGS

# --- Conditional Processing Block ---
if (file.exists(fp_boundary_wftdm)) {
  # File exists, skip processing
  message("✅ Analysis boundary already exists: ", fp_boundary_wftdm)
} else {
  # File does NOT exist, proceed with DuckDB SQL processing

  # Initialize In-memory database
  conn <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(conn, shutdown = TRUE))

  # install and load spatial extension
  DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
  DBI::dbExecute(conn, "SET geometry_always_xy = true")

  # Process using DuckDB SQL
  query_prep <- glue::glue(
    "
    COPY (
      WITH raw_union AS (
        -- 1. Initial Dissolve
        SELECT ST_Union_Agg(geometry) as geom
        FROM read_parquet('{fp_regional_bound}')
        WHERE plan_org IN ('MAG MPO', 'WFRC MPO')
      ),

      exploded_parts AS (
        -- 2. Handle MultiPolygons
        SELECT unnest(ST_Dump(geom)) as d
        FROM raw_union
      ),

      filled_polys AS (
        -- 3. Strip Holes (Make Exterior Ring -> Polygon)
        SELECT ST_MakePolygon(ST_ExteriorRing(d.geom)) as geom
        FROM exploded_parts
      ),

      final_union AS (
        -- 4. Re-Aggregate
        SELECT ST_Union_Agg(geom) as geom
        FROM filled_polys
      )

      -- 5. FINAL SELECT: Explicit Transform
      --    ST_Transform(geometry, source_crs, target_crs, always_xy := true)
      SELECT
        ST_AsWKB(
          ST_Transform(geom, '{SOURCE_CRS}', 'EPSG:4326', always_xy := true)
        ) as geometry
      FROM final_union

    ) TO '{fp_boundary_wftdm}' (FORMAT PARQUET)
  "
  )

  # Assuming 'conn' is your active DuckDB connection
  DBI::dbExecute(conn, query_prep)
  message("✅ Analysis boundary created (DuckDB): ", fp_boundary_wftdm)
}
```

```{r}
#| eval: false

arrow::open_dataset(fp_boundary_wftdm) |>
  sf::st_as_sf(crs = CRS_WGS) |>
  mapgl::maplibre_view()
```

## Generate H3 Hexagons

```r
#| label: gen-h3-grid

# -----------------------------------------------------------------------------
# DOESN'T WORK CURRENTLY BECAUSE {duckspatial} IS STILL UNDER DEVELOPMENT
# -----------------------------------------------------------------------------

fp_h3_grid <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")
H3_RES     <- 11

if (file.exists(fp_h3_grid)) {
  message("✅ H3 Grid already exists: ", fp_h3_grid)
} else {

  conn <- duckspatial::ddbs_create_conn()
  on.exit(duckspatial::ddbs_stop_conn(conn), add = TRUE)

  duckspatial::ddbs_load(conn)
  DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
  DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")

  # Define Lazy Table from the cleaned boundary
  tbl_clean_boundary <- dplyr::tbl(conn, paste0("read_parquet('", fp_boundary_wftdm, "')"))

  h3_grid <- tbl_clean_boundary |>

    # A. Dump MultiPolygons (Using hypothetical wrapper)
    # MISSING FUNCTION: ddbs_dump()
    duckplyr::mutate(parts = ddbs_dump(geometry)) |>
    tidyr::unnest_longer(parts) |>

    # B. Convert to WKT (Using hypothetical wrapper)
    # MISSING FUNCTION: ddbs_as_text()
    duckplyr::mutate(wkt_geom = ddbs_as_text(parts$geom)) |>

    # C. H3 Polyfill (Manual SQL as requested)
    # DuckDB's H3 extension functions are not standard OGC, so we inject SQL
    duckplyr::mutate(
      h3_indices = dbplyr::sql(glue::glue("h3_polygon_wkt_to_cells(wkt_geom, {H3_RES})"))
    ) |>
    tidyr::unnest_longer(h3_indices) |>

    # D. Convert H3 Index back to Geometry (Manual SQL)
    duckplyr::transmute(
      h3_index = h3_indices,
      geometry = dbplyr::sql("h3_cell_to_boundary_wkb(h3_indices)")
    ) |>

    duckplyr::distinct()

  # Write Result
  query_h3_sql <- dbplyr::sql_render(h3_grid)
  DBI::dbExecute(conn, glue::glue("COPY ({query_h3_sql}) TO '{fp_h3_grid}' (FORMAT PARQUET)"))

  message("✅ H3 Grid saved to: ", fp_h3_grid)
}
```

```{r}
#| label: gen-h3-grid

# 1. Define Inputs/Outputs
fp_boundary_wftdm <- here::here(
  sub_dirs$base,
  "Analysis_Boundary_WFRC_MAG.parquet"
)
fp_h3_grid <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")
H3_RES <- 10 # Reference: https://gist.github.com/colbyn/001064f00385d253b42693c3889f9beb

# The input boundary should now be EPSG:4326 from the previous step.
# We define it explicitly here for the 3-argument ST_Transform.
BOUNDARY_CRS <- CRS_WGS

if (file.exists(fp_h3_grid)) {
  message("✅ H3 Grid already exists: ", fp_h3_grid)
} else {
  conn <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(conn, shutdown = TRUE))

  DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
  DBI::dbExecute(conn, "SET geometry_always_xy = true")
  DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")

  # Fix: Removed curly braces from comments to prevent glue parsing errors
  query_h3 <- glue::glue(
    "
    COPY (
      WITH boundary_parts AS (
        -- 1. Read Boundary & FORCE Transform to 4326
        --    ST_Transform(geom, source_crs, target_crs, always_xy := true)
        SELECT
          unnest(
            ST_Dump(
              ST_Transform(geometry, '{BOUNDARY_CRS}', 'EPSG:4326', always_xy := true)
            )
          ) as d
        FROM read_parquet('{fp_boundary_wftdm}')
      ),

      h3_indices AS (
        SELECT
          -- 2. Generate H3 Indices (Polyfill)
          --    ST_AsText creates the WKT needed by H3
          unnest(h3_polygon_wkt_to_cells(ST_AsText(d.geom), {H3_RES})) as h3_index
        FROM boundary_parts
      )

      SELECT DISTINCT
        h3_index,
        -- 3. Convert Index back to Geometry
        --    H3 boundaries are naturally 4326
        h3_cell_to_boundary_wkb(h3_index) as geometry
      FROM h3_indices

    ) TO '{fp_h3_grid}' (FORMAT PARQUET)
  "
  )

  tryCatch(
    {
      DBI::dbExecute(conn, query_h3)
      message("✅ H3 Grid saved to: ", fp_h3_grid)
    },
    error = function(e) {
      stop("❌ H3 Generation Failed: ", e$message)
    }
  )
}
```

```{r}
#| eval: false

arrow::read_parquet(fp_h3_grid) |>
  sf::st_as_sf(crs = CRS_WGS) |>
  mapgl::maplibre_view()
```

## Set Isochrone Inputs and Configuration

```{r}
#| label: config-layers

# Master Input Configuration
# - name: Output filename prefix
# - filter: SQL filtering logic (default "1=1")
# - paths: List of file paths (if >1, they are unioned)
layer_config <- tibble::tribble(
  ~name               , ~filter                                       , ~path                                                                         ,

  # --- TRANSIT (Split by Mode) ---
  "Transit_Local"     , "Mode IN ('Bus', 'Flex Bus', 'Microtransit')" , here::here(sub_dirs$transport, "UTA_Stops_and_Most_Recent_Ridership.parquet") ,
  "Transit_Rapid"     , "Mode IN ('BRT', 'TRAX', 'BRT, Bus')"         , here::here(sub_dirs$transport, "UTA_Stops_and_Most_Recent_Ridership.parquet") ,
  "Transit_CRT"       , "Mode = 'FrontRunner'"                        , here::here(sub_dirs$transport, "UTA_Stops_and_Most_Recent_Ridership.parquet") ,

  # --- INFRASTRUCTURE ---
  "Freeways"          , "1=1"                                         , here::here(sub_dirs$transport, "Freeway_Exit_Locations.parquet")              ,

  # --- AMENITIES ---
  "ChildCare"         , "1=1"                                         , here::here(sub_dirs$amenities, "ChildCareFacilities.parquet")                 ,
  "Healthcare"        , "1=1"                                         , here::here(sub_dirs$amenities, "LicensedHealthCareFacilities.parquet")        ,
  "Schools_PreK12"    , "1=1"                                         , here::here(sub_dirs$amenities, "Schools_PreKto12.parquet")                    ,
  "Schools_HigherEd"  , "1=1"                                         , here::here(sub_dirs$amenities, "Schools_HigherEducation.parquet")             ,
  "Grocery"           , "1=1"                                         , here::here(sub_dirs$amenities, "UtahGroceryAndFoodStores_DAF.parquet")        ,
  "Community_Centers" , "1=1"                                         , here::here(sub_dirs$amenities, "Community_Centers.parquet")
)
```

## Generate Walking Isochrones

```{r}
#| label: iso-walking

layer_config |>
  # Optional: You can filter the config first if you only want specific layers for driving
  # dplyr::filter(name %in% c("Grocery", "Healthcare")) |>
  purrr::pwalk(function(name, filter, path) {
    # Direct execution (No union logic needed)
    process_isochrones(
      path = path,
      name = name,
      profile = "foot-walking",
      range = c(300, 600, 900, 1800, 3599),
      input_filter = filter,
      boundary_path = fp_boundary_wftdm,
      save_dir = iso_dirs$walking,
      dry_run = TRUE # Set to FALSE to generate files
    )
  })
```

## Generate Biking Isochrones

```{r}
#| label: iso-biking

layer_config |>
  # Optional: You can filter the config first if you only want specific layers for driving
  # dplyr::filter(name %in% c("Grocery", "Healthcare")) |>
  purrr::pwalk(function(name, filter, path) {
    process_isochrones(
      path = path,
      name = name,
      profile = "cycling-regular",
      range = c(300, 600, 900, 1800, 3599),
      input_filter = filter,
      boundary_path = fp_boundary_wftdm,
      save_dir = iso_dirs$biking,
      dry_run = TRUE # Set to FALSE to generate files
    )
  })
```

## Generate Driving Isochrones

```{r}
#| label: iso-driving

layer_config |>
  # Optional: You can filter the config first if you only want specific layers for driving
  # dplyr::filter(name %in% c("Grocery", "Healthcare")) |>
  purrr::pwalk(function(name, filter, path) {
    process_isochrones(
      path = path,
      name = name,
      profile = "driving-car",
      range = c(300, 600, 900, 1800, 3599),
      input_filter = filter,
      boundary_path = fp_boundary_wftdm,
      save_dir = iso_dirs$driving,
      dry_run = TRUE # Set to FALSE to generate files
    )
  })
```

## Graded Park Access Buffers

```{r}
#| label: process-parks-graded

# 1. Define Inputs/Outputs
fp_park_5_in <- here::here(sub_dirs$amenities, "Parks_5min_ServiceArea.parquet")
fp_park_10_in <- here::here(
  sub_dirs$amenities,
  "Parks_10min_ServiceArea.parquet"
)
fp_parks_graded <- here::here(sub_dirs$amenities, "ParkAccess_Graded.parquet")

# 2. Define Buffer Parameters
#    Base Speed: 3 MPH (~0.25 miles per 5 mins).
#    If you strictly meant 0.25 MPH, change this value to 0.25.
SPEED_MPH <- 3.0
MPH_TO_MPS <- 0.44704 # Conversion factor: Miles/Hr to Meters/Sec

# Calculate distances in meters based on time delta from the 10-min (600s) layer
# Distance = (Target_Time - Base_Time) * Speed
calc_buffer <- function(target_seconds) {
  seconds_delta <- target_seconds - 600
  return(seconds_delta * SPEED_MPH * MPH_TO_MPS)
}

dist_900 <- calc_buffer(900) # ~ 15 min total
dist_1800 <- calc_buffer(1800) # ~ 30 min total
dist_3600 <- calc_buffer(3599) # ~ 60 min total

# 3. Conditional Processing
if (file.exists(fp_parks_graded)) {
  message("✅ Graded Park Access already exists: ", fp_parks_graded)
} else {
  # Initialize DB
  conn <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(conn, shutdown = TRUE))
  DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")

  # SQL Logic:
  # 1. Read 5min (Val 300) and 10min (Val 600)
  # 2. Buffer 10min to create 900, 1800, 3599 tiers
  # 3. Union and Save
  query <- glue::glue(
    "
    COPY (
      WITH
      -- 1. Load 5-Min Layer (Inner Core)
      layer_5min AS (
        SELECT 300 as value,
               ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
        FROM read_parquet('{fp_park_5_in}')
      ),

      -- 2. Load 10-Min Layer (Base for Buffering)
      layer_10min AS (
        SELECT 600 as value,
               ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
        FROM read_parquet('{fp_park_10_in}')
      ),

      -- 3. Generate All Tiers
      tiers AS (
        -- Tier: 0-5 min (Provided)
        SELECT value, geom FROM layer_5min

        UNION ALL

        -- Tier: 5-10 min (Provided)
        SELECT value, geom FROM layer_10min

        UNION ALL

        -- Tier: 15 min (900s) -> Buffer 10min by ~0.25 miles
        SELECT 900 as value, ST_Buffer(geom, {dist_900}) as geom
        FROM layer_10min

        UNION ALL

        -- Tier: 30 min (1800s) -> Buffer 10min by ~1.0 miles
        SELECT 1800 as value, ST_Buffer(geom, {dist_1800}) as geom
        FROM layer_10min

        UNION ALL

        -- Tier: 60 min (3599s) -> Buffer 10min by ~2.5 miles
        SELECT 3599 as value, ST_Buffer(geom, {dist_3600}) as geom
        FROM layer_10min
      )

      -- 4. Final Transform back to WGS84
      SELECT
        value,
        ST_Transform(geom, '{CRS_UTM}', '{CRS_WGS}', always_xy := true) as geometry
      FROM tiers
      WHERE NOT ST_IsEmpty(geom)

    ) TO '{fp_parks_graded}' (FORMAT PARQUET)
  "
  )

  DBI::dbExecute(conn, query)
  message("✅ Saved Graded Park Access to: ", fp_parks_graded)
}
```

## Verify Results

```{r}
#| label: verify-isochrones
#| eval: false

# 1. Verify Folder Structure
fs::dir_tree(dirs$isochrones)
```


```{r}
#| eval: false

# 2. Visual Check: Compare Walking vs Biking vs Driving
preview_layer(
  here::here(iso_dirs$walking, "Schools_HigherEd.parquet"),
  column = "value"
)
```

```{r}
#| eval: false

# Biking - Should be larger
preview_layer(
  here::here(iso_dirs$biking, "Schools_HigherEd.parquet"),
  column = "value"
)
```

```{r}
#| eval: false

# Driving Should be much larger
preview_layer(
  here::here(iso_dirs$driving, "Schools_HigherEd.parquet"),
  column = "value"
)
```
