---
title: Process Data Layers
subtitle: Lorem Ipsum
description: Process the data to prepare for spatial operations (buffer, isochrones, etc.)
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here"         # https://github.com/r-lib/here/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Setup OpenRouteService API

```{r}
#| label: setup-ors

# Ensure your _environment file has: ORS_API_KEY="your_key_here"
readRenviron(here::here("_environment"))

# 2. Check Key
if (Sys.getenv("ORS_API_KEY") == "") {
  stop("❌ ORS_API_KEY not found in _environment file.")
} else {
  # Set it for the package
  ors_api_key(Sys.getenv("ORS_API_KEY"))
  message("✅ API Key Loaded")
}
```

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source = here::here("_src"), # Source custom functions
  raw = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results = here::here("_output") # Final Output: Analysis Results
)
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base   = here::here(dirs$processed, "base_layers"),
  parcels = here::here(dirs$processed, "parcels_taz"),
  landuse = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "process_isochrones.R")) # Generate isochrones using OpenRouteService API
source(here::here(dirs$source, "create_buffer.R")) # Generate equidistant buffers
source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

# Preprocessing

## Set Regional Boundary

```{r}
#| label: prep-boundary

# Define inputs/outputs
fp_regional_bound <- here::here(sub_dirs$base, "RegionalBoundaryComponents.parquet")
fp_boundary_wftdm <- here::here(sub_dirs$base, "Analysis_Boundary_WFRC_MAG.parquet")

# --- Conditional Processing Block ---
if (file.exists(fp_boundary_wftdm)) {
  # File exists, skip processing
  message("✅ Analysis boundary already exists: ", fp_boundary_wftdm)
} else {
  # File does NOT exist, proceed with DuckDB SQL processing

  # Initialize In-memory database
  conn <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(conn, shutdown = TRUE))

  # install and load spatial extension
  DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
  DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")

  # Process using DuckDB SQL
  query_prep <- glue::glue("
    COPY (
      WITH raw_union AS (
        -- 1. Initial Dissolve (Result likely has holes/slivers)
        SELECT ST_Union_Agg(geometry) as geom
        FROM read_parquet('{fp_regional_bound}')
        WHERE plan_org IN ('MAG MPO', 'WFRC MPO')
      ),

      exploded_parts AS (
        -- 2. Handle MultiPolygons: Dump into individual polygons
        --    We alias the struct output as 'd' to access d.geom
        SELECT unnest(ST_Dump(geom)) as d
        FROM raw_union
      ),

      filled_polys AS (
        -- 3. Strip Holes:
        --    Get Exterior Ring (LineString) -> Make new Polygon (No internal rings)
        SELECT ST_MakePolygon(ST_ExteriorRing(d.geom)) as geom
        FROM exploded_parts
      )

      -- 4. Re-Aggregate (in case there were legitimate islands) & Save
      SELECT ST_AsWKB(ST_Union_Agg(geom)) as geometry
      FROM filled_polys

    ) TO '{fp_boundary_wftdm}' (FORMAT PARQUET)
  ")

  # Assuming 'conn' is your active DuckDB connection
  DBI::dbExecute(conn, query_prep)
  message("✅ Analysis boundary created (DuckDB): ", fp_boundary_wftdm)
}
```

```{r}
#| eval: false

arrow::open_dataset(fp_boundary_wftdm) |>
  sf::st_as_sf(crs = CRS_WGS) |>
  mapgl::maplibre_view()
```

## Generate H3 Hexagons

```{r}
#| label: gen-h3-grid

# 1. Define Inputs/Outputs
fp_boundary_wftdm <- here::here(sub_dirs$base, "Analysis_Boundary_WFRC_MAG.parquet")
fp_h3_grid        <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")
H3_RES            <- 7 # Reference: https://gist.github.com/colbyn/001064f00385d253b42693c3889f9beb

if (file.exists(fp_h3_grid)) {
  message("✅ H3 Grid already exists: ", fp_h3_grid)
} else {

  conn <- DBI::dbConnect(duckdb::duckdb())
  on.exit(DBI::dbDisconnect(conn, shutdown = TRUE))

  DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
  DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
  DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")

  # Fix: Removed curly braces from comments to prevent glue parsing errors
  query_h3 <- glue::glue("
    COPY (
      WITH boundary_parts AS (
        SELECT
          -- 1. Dump MultiPolygon to list of structs [geom, path]
          -- 2. Unnest to rows. We alias the struct as 'd'
          unnest(ST_Dump(geometry)) as d
        FROM read_parquet('{fp_boundary_wftdm}')
      ),

      h3_indices AS (
        SELECT
          -- 3. Access the '.geom' field of the struct 'd'
          -- 4. Convert to WKT (Text) for H3 Polyfill
          unnest(h3_polygon_wkt_to_cells(ST_AsText(d.geom), {H3_RES})) as h3_index
        FROM boundary_parts
      )

      SELECT DISTINCT
        h3_index,
        -- 5. Convert Cell ID back to WKB Geometry
        h3_cell_to_boundary_wkb(h3_index) as geometry
      FROM h3_indices

    ) TO '{fp_h3_grid}' (FORMAT PARQUET)
  ")

  tryCatch({
    DBI::dbExecute(conn, query_h3)
    message("✅ H3 Grid saved to: ", fp_h3_grid)
  }, error = function(e) {
    stop("❌ H3 Generation Failed: ", e$message)
  })
}
```

```{r}
#| eval: false

arrow::read_parquet(fp_h3_grid) |>
  sf::st_as_sf(crs = CRS_WGS) |>
  mapgl::maplibre_view()
```

## Set Isochrone Inputs and Configuration

```{r}
#| label: isochrone-setup

# We use the parquet files created in step 1
input_layers <- c(
  here::here(sub_dirs$transport, "UTA_Stops_and_Most_Recent_Ridership.parquet"), # Transit (All Modes)
  here::here(sub_dirs$amenities, "Community_Centers.parquet"),
  here::here(sub_dirs$amenities, "ChildCareFacilities.parquet"),
  here::here(sub_dirs$amenities, "UtahGroceryAndFoodStores_DAF.parquet"),
  here::here(sub_dirs$amenities, "LicensedHealthCareFacilities.parquet"),
  here::here(sub_dirs$amenities, "Schools_PreKto12.parquet"),
  here::here(sub_dirs$amenities, "Schools_HigherEducation.parquet")
)
```

```{r}
#| label: isochrone-batch-helper

# Batch Helper Function
# Takes a list of files -> Generates Isochrones using original filenames
run_batch <- function(file_list, profile, range, ...) {

  purrr::walk(file_list, function(input_path) {

    # 1. Get original name (e.g. "ChildCareFacilities")
    original_name <- fs::path_ext_remove(fs::path_file(input_path))

    # 2. Execute
    # process_isochrones automatically handles the subfolder creation based on 'profile'
    process_isochrones(
      path        = input_path,
      name        = original_name,  # Output name matches input name
      profile     = profile,
      range       = range,
      boundary_path = fp_boundary_wftdm,
      src_crs       = CRS_STP,
      save_dir    = dirs$isochrones,
      dry_run     = TRUE,          # Set to TRUE to test
      batch_size  = 5,
      ...                           # Pass extra args
    )
  })
}
```

## Generate Walking Isochrones

```{r}
#| label: isochrone-walking

run_batch(
  file_list = input_layers,
  profile   = "foot-walking",
  range     = c(300, 600, 900)
)
```

## Generate Biking Isochrones

```{r}
#| label: isochrone-biking

run_batch(
  file_list = input_layers,
  profile   = "cycling-regular",
  range     = c(300, 600, 900)
)
```

## Generate Driving Isochrones

```{r}
#| label: isochrone-driving

run_batch(
  file_list = input_layers,
  profile   = "driving-car",
  range     = c(300, 600, 900)
)
```

## Verify Results

```{r}
#| label: verify-isochrones
#| eval: false

# 1. Verify Folder Structure
fs::dir_tree(dirs$isochrones)
```


```{r}
#| eval: false

# 2. Visual Check: Compare Walking vs Driving for Grocery
# Walking (Blue)
preview_layer(
  here::here(dirs$isochrones, "foot-walking", "UtahGroceryAndFoodStores_DAF.parquet"),
  color = "blue",
  limit = 100
)
```


```{r}
#| eval: false

# Driving (Red) - Should be much larger
preview_layer(
  here::here(dirs$isochrones, "driving-car", "UtahGroceryAndFoodStores_DAF.parquet"),
  color = "red",
  limit = 100
)
```
