---
title: Spatial Operations
subtitle: Enrich Parcels and H3 Grids with Amenity Access
description: Perform spatial joins between target geometries (Parcels, H3) and isochrone layers to calculate access times.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-12"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here"         # https://github.com/r-lib/here/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source     = here::here("_src"), # Source custom functions
  raw        = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed  = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered   = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results    = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base      = here::here(dirs$processed, "base_layers"),
  parcels   = here::here(dirs$processed, "parcels_taz"),
  landuse   = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access    = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-iso-subdirs

# Define Sub-directories for Isochrones
iso_dirs <- list(
  walking   = here::here(dirs$isochrones, "foot-walking"),
  biking    = here::here(dirs$isochrones, "cycling-regular"),
  driving   = here::here(dirs$isochrones, "driving-car")
)

purrr::walk(iso_dirs, ~fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

## Initialize Database

We use DuckDB for all spatial operations to avoid loading millions of geometries into R memory.

```{r}
#| label: init-db

# Create persistent connection for this session
conn <- DBI::dbConnect(duckdb::duckdb())

# Load Spatial Extensions
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
```

# Part 1: Prepare Targets (The Canvas)

We start by loading the Parcels and H3 grids. We immediately join Administrative and Building attributes so our "Canvas" is fully described before we start painting it with amenities.

## Load H3 Grid

We load H3 first to get the resolution, then use that resolution to load Parcels with the correct H3 index pre-calculated.

```{r}
#| label: load-h3

fp_h3_grid <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")

# 1. Load Polygons (UTM)
#    Note: H3 WKB is usually WGS84, so we must transform to UTM.
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_h3 AS
  SELECT
    CAST(h3_index AS UBIGINT) as h3_index, -- Strict type safety
    ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
  FROM read_parquet('{fp_h3_grid}')
"))
DBI::dbExecute(conn, "CREATE INDEX idx_h3_geom ON targets_h3 USING RTREE (geom)")

# 2. OPTIMIZATION: Create & Index Points
message("📍 Pre-calculating H3 Points (PointOnSurface)...")
DBI::dbExecute(conn, "
  CREATE OR REPLACE TABLE points_h3 AS
  SELECT rowid, h3_index, ST_Centroid(geom) as geom
  FROM targets_h3
")
DBI::dbExecute(conn, "CREATE INDEX idx_h3_utm ON points_h3 USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_h3_id ON points_h3 (h3_index)")

message("✅ H3 Grid loaded & indexed")
```

```{r}
# -----------------------------------------------------------------------------
# 2. GET RESOLUTION
# -----------------------------------------------------------------------------
# We need this 'res' variable for the next steps
res <- DBI::dbGetQuery(conn, "SELECT h3_get_resolution(h3_index) as res FROM targets_h3 LIMIT 1")$res
message(glue::glue("ℹ️ Detected Grid Resolution: {res}"))
```

## Load Parcels

```{r}
#| label: load-parcels

fp_remm_parcels <- here::here(sub_dirs$parcels, "REMM_BY_parcels.parquet")

# 1. Load Polygons (UTM)
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_parcels AS
  SELECT
    parcel_id,

    -- 1. Calculate H3 Index DIRECTLY from raw WGS84 Centroid
    CAST(h3_latlng_to_cell(
      ST_Y(ST_Centroid(geometry)),
      ST_X(ST_Centroid(geometry)),
      {res}
    ) AS UBIGINT) as h3_index,

    -- 2. Transform to UTM for storage/buffer ops
    ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)) as geom
  FROM read_parquet('{fp_remm_parcels}')
  WHERE ST_IsValid(geometry)
"))

# 2. OPTIMIZATION: Create & Index Points (Point-On-Surface)
#    ST_PointOnSurface guarantees the point is INSIDE the polygon (unlike Centroid).
#    This handles U-shaped parcels correctly.
DBI::dbExecute(conn, "
  CREATE OR REPLACE TABLE points_parcels AS
  SELECT rowid, h3_index, ST_PointOnSurface(geom) as geom
  FROM targets_parcels
")
DBI::dbExecute(conn, "CREATE INDEX idx_pp_utm ON points_parcels USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_pp_h3 ON points_parcels (h3_index)")

message("✅ Parcels loaded & indexed")
```

# Part 2: Static Attributes (Admin & Land Use)

## Attach Administrative Data

A helper function to join City/County codes to any target table (Parcel or H3).

```{r}
#| label: func-admin-join

attach_admin_data <- function(conn, target_table, point_table) {

  message(glue::glue("🏛️ Attaching Admin Data to {target_table}..."))

  # 1. Municipalities (CommCode)
  fp_muni <- here::here(sub_dirs$base, "UtahMunicipalBoundaries.parquet")

  # Create temp layer in UTM & Index
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_muni AS
    SELECT
      ugrcode as CommCode,
      ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_muni}')
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_muni ON layer_muni USING RTREE (geom)")

  # Spatial Join using Pre-Calculated Points
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS CommCode VARCHAR;

    UPDATE {target_table}
    SET CommCode = matched.CommCode
    FROM (
      SELECT
        tp.rowid,
        m.CommCode
      FROM {point_table} tp, layer_muni m
      WHERE ST_Intersects(tp.geom, m.geom)
    ) matched
    WHERE {target_table}.rowid = matched.rowid;
  "))

  # 2. Counties (COUNTYNBR)
  fp_county <- here::here(sub_dirs$base, "UtahCountyBoundaries.parquet")

  # Create temp layer in UTM & Index
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_county AS
    SELECT
      countynbr as COUNTYNBR,
      ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_county}')
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_county ON layer_county USING RTREE (geom)")

  # Spatial Join using Pre-Calculated Points
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS COUNTYNBR VARCHAR;

    UPDATE {target_table}
    SET COUNTYNBR = matched.COUNTYNBR
    FROM (
      SELECT
        tp.rowid,
        c.COUNTYNBR
      FROM {point_table} tp, layer_county c
      WHERE ST_Intersects(tp.geom, c.geom)
    ) matched
    WHERE {target_table}.rowid = matched.rowid;
  "))

  # Cleanup
  DBI::dbExecute(conn, "DROP TABLE layer_muni; DROP TABLE layer_county;")
}
```

```{r}
#| label: func-admin-join

attach_admin_data <- function(conn, target_table, point_table) {

  message(glue::glue("🏛️ Attaching Admin Data to {target_table}..."))

  # 1. Municipalities (CommCode)
  fp_municipal_bound <- here::here(sub_dirs$base, "UtahMunicipalBoundaries.parquet")

  # OPTIMIZATION: Simplify Geometry by 1.0 meters
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_muni AS
    SELECT
      ugrcode as CommCode,
      ST_Simplify(ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)), 1.0) as geom
    FROM read_parquet('{fp_municipal_bound}')
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_muni ON layer_muni USING RTREE (geom)")

  # Spatial Join
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS CommCode VARCHAR;

    UPDATE {target_table}
    SET CommCode = matched.CommCode
    FROM (
      SELECT tp.rowid, m.CommCode
      FROM {point_table} tp, layer_muni m
      WHERE ST_Intersects(tp.geom, m.geom)
    ) matched
    WHERE {target_table}.rowid = matched.rowid;
  "))

  # 2. Counties (COUNTYNBR)
  fp_county_bound <- here::here(sub_dirs$base, "UtahCountyBoundaries.parquet")

  # OPTIMIZATION: Simplify Geometry by 1.0 meters
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_county AS
    SELECT
      countynbr as COUNTYNBR,
      ST_Simplify(ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)), 1.0) as geom
    FROM read_parquet('{fp_county_bound}')
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_county ON layer_county USING RTREE (geom)")

  # Spatial Join
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS COUNTYNBR VARCHAR;

    UPDATE {target_table}
    SET COUNTYNBR = matched.COUNTYNBR
    FROM (
      SELECT tp.rowid, c.COUNTYNBR
      FROM {point_table} tp, layer_county c
      WHERE ST_Intersects(tp.geom, c.geom)
    ) matched
    WHERE {target_table}.rowid = matched.rowid;
  "))

  # Cleanup
  DBI::dbExecute(conn, "DROP TABLE layer_muni; DROP TABLE layer_county;")
}
```

```{r}
#| label: run-admin-join

# Pass the pre-calculated point tables we created in Part 1
attach_admin_data(conn, "targets_parcels", "points_parcels")
attach_admin_data(conn, "targets_h3", "points_h3")
```

## Building Codes (BC)

We join the REMM buildings table. Since `building_type` contains full strings (e.g., "Single Family Res"), we create a mapping table to translate them to abbreviations (e.g., "SF").

```{r}
#| label: building-codes

# -----------------------------------------------------------------------------
# 1. Load Buildings & Store as UTM (Projected)
# -----------------------------------------------------------------------------
message("🏗️ Loading Building Centroids (Spatial)...")

fp_remm_buildings <- here::here(sub_dirs$parcels, "REMM_BY_buildings.parquet")

# Create Temp Table with BOTH UTM Geom and H3 Index
# Load -> Centroid -> Transform (UTM) -> Index
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TEMP TABLE points_buildings AS
  SELECT
    parcel_id,
    building_type,
    building_sqft,

    -- 1. H3 Index (Integer Match)
    CAST(h3_latlng_to_cell(
      ST_Y(ST_Centroid(geometry)),
      ST_X(ST_Centroid(geometry)),
      {res}
    ) AS UBIGINT) as h3_index,

    -- 2. UTM Geometry (Spatial Match)
    ST_Transform(ST_Centroid(geometry), '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom

  FROM read_parquet('{fp_remm_buildings}')
  WHERE ST_IsValid(geometry)
"))

# Index both columns for speed
DBI::dbExecute(conn, "CREATE INDEX idx_b_utm ON points_buildings USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_b_h3 ON points_buildings (h3_index)")

# Create Mapping Table (Full Name -> Abbreviation)
bc_mapping <- tibble::tribble(
  ~full_name,                  ~abbr,
  "Agriculture",              "AG",
  "Church",                   "CH",
  "Commons",                  "OS",
  "Educational",              "ED",
  "Empty - Buildable",        "EM",
  "Empty Buildable",          "EM",
  "Governmental",             "GO",
  "Group Quarters",           "GQ",
  "Healthcare",               "HE",
  "Industrial",               "IN",
  "Mixed Use",                "MU",
  "Multi Family Res",         "MF",
  "No Build",                 "NB",
  "Office",                   "OF",
  "Open Space Non-Buildable", "OS",
  "Other",                    "OT",
  "Retail",                   "RE",
  "Single Family Res",        "SF",
  "Utilities",                "UT"
  # Note: NA values are handled in the join below
)
DBI::dbWriteTable(conn, "bc_map", bc_mapping, overwrite = TRUE)
```

```{r}
# -----------------------------------------------------------------------------
# 2. PARCELS: Attribute Join (Parcel ID Match)
# -----------------------------------------------------------------------------
message("🏗️ Attaching Building Codes to Parcels (Attribute Match)...")

# Join Abbreviation to Parcels
# Note: QUALIFY ROW_NUMBER() handles cases where a parcel has multiple buildings
# Note: We prioritize the LARGEST building on the parcel using building_sqft
DBI::dbExecute(conn, "
  ALTER TABLE targets_parcels ADD COLUMN BC VARCHAR;

  UPDATE targets_parcels
  SET BC = COALESCE(mapped.abbr, 'OT') -- Handle NA/Unknown as 'OT'
  FROM (
    SELECT
      b.parcel_id,
      m.abbr
    FROM points_buildings b
    JOIN bc_map m ON b.building_type = m.full_name

    -- Order by square footage descending to pick the main building
    QUALIFY ROW_NUMBER() OVER (
      PARTITION BY b.parcel_id
      ORDER BY b.building_sqft DESC
    ) = 1
  ) mapped
  WHERE targets_parcels.parcel_id = mapped.parcel_id
")
```

```{r}
# -----------------------------------------------------------------------------
# 3. H3: Aggregation via Index Match (Math-Based)
# -----------------------------------------------------------------------------
message("🏗️ Aggregating Building Codes to H3 (Index Match)...")

DBI::dbExecute(conn, glue::glue("
  ALTER TABLE targets_h3 ADD COLUMN IF NOT EXISTS BC VARCHAR;

  UPDATE targets_h3
  SET BC = agg.mode_bc
  FROM (
    SELECT
      b.h3_index,
      -- COALESCE only works if the row exists, so LEFT JOIN is required
      MODE(COALESCE(m.abbr, 'OT')) as mode_bc
    FROM points_buildings b
    LEFT JOIN bc_map m ON b.building_type = m.full_name
    GROUP BY b.h3_index
  ) agg
  WHERE targets_h3.h3_index = agg.h3_index
"))

# Cleanup
DBI::dbExecute(conn, "DROP TABLE points_buildings; DROP TABLE bc_map;")
```

## Attach Job Access (AA, AT)

We attach the raw job access numbers from the TAZ layer. The normalization (dividing by city max) will happen in Step 4.

```{r}
#| label: func-job-join

attach_job_access <- function(conn, target_table, point_table) {

  message(glue::glue("💼 Attaching Job Access to {target_table}..."))
  fp_taz_access <- here::here(dirs$processed, "access", "TAZ_AccessToOpportunities_OY2023.parquet")

  # 1. Load TAZ & Index
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE taz_access AS
    SELECT
      ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom,
      job_by_auto,
      job_by_tran
    FROM read_parquet('{fp_taz_access}')
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_taz_geom ON taz_access USING RTREE (geom)")

  # 2. Spatial Join using Pre-Calculated Points
  # Note: "AA" and "AT" must be quoted because they can be reserved keywords
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS \"AA\" DOUBLE;
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS \"AT\" DOUBLE;

    UPDATE {target_table}
    SET \"AA\" = match.job_by_auto, \"AT\" = match.job_by_tran
    FROM (
      SELECT tp.rowid, t.job_by_auto, t.job_by_tran
      FROM {point_table} tp, taz_access t
      WHERE ST_Intersects(tp.geom, t.geom)
    ) match
    WHERE {target_table}.rowid = match.rowid;
  "))

  DBI::dbExecute(conn, "DROP TABLE taz_access;")
}
```

```{r}
#| label: run-job-join

# Run with pre-calculated points
attach_job_access(conn, "targets_parcels", "points_parcels")
attach_job_access(conn, "targets_h3", "points_h3")
```

```{r}
print(DBI::dbGetQuery(conn, "SELECT COUNT(*) as Parcels_With_Jobs FROM targets_parcels WHERE \"AA\" IS NOT NULL"))
```

# Part 3: Amenity Operations Engine

## Configuration

We define all external layers here. This structure handles Binary (In/Out), Isochrone (Time), and Buffer (Distance) logic.

```{r}
#| label: config-amenities

# Helpers
iso_walk  <- function(f) here::here(dirs$isochrones, "foot-walking", f)
iso_drive <- function(f) here::here(dirs$isochrones, "driving-car", f)
land_use  <- function(f) here::here(sub_dirs$landuse, f)
trans     <- function(f) here::here(sub_dirs$transport, f)
amen      <- function(f) here::here(sub_dirs$amenities, f)

amenity_config <- tibble::tribble(
  ~col_name, ~type,       ~path,                                    ~filter_sql,               ~value_col, ~buffer_opts,

  # --- BINARY ---
  "OZ",      "binary",    land_use("Utah_Qualified_Opportunity_Zones.parquet"), "1=1",              NA,         NA,
  "AP",      "binary",    amen("WalkAccesstoParks_10Min.parquet"),              "1=1",              NA,         NA,
  "CM",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Metropolitan Center'", NA, NA,
  "CU",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Urban Center'",        NA, NA,
  "CC",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'City Center'",         NA, NA,
  "CN",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Neighborhood Center'", NA, NA,

  # --- ISOCHRONES ---
  "TT_Local", "isochrone", iso_walk("Transit_Local.parquet"),       "1=1", "value", NA,
  "TT_Rapid", "isochrone", iso_walk("Transit_Rapid.parquet"),       "1=1", "value", NA,
  "TT_CRT",   "isochrone", iso_walk("Transit_CRT.parquet"),         "1=1", "value", NA,
  "TF",      "isochrone", iso_drive("Freeways.parquet"),            "1=1", "value", NA,
  "AC",      "isochrone", iso_walk("ChildCare.parquet"),            "1=1", "value", NA,
  "AH",      "isochrone", iso_drive("Healthcare.parquet"),          "1=1", "value", NA,
  "AG",      "isochrone", iso_walk("Grocery.parquet"),              "1=1", "value", NA,
  "AM",      "isochrone", iso_walk("Community_Centers.parquet"),    "1=1", "value", NA,
  "AE_K12",  "isochrone", iso_walk("Schools_PreK12.parquet"),       "1=1", "value", NA,
  "AE_High", "isochrone", iso_drive("Schools_HigherEd.parquet"),    "1=1", "value", NA,

  # --- BUFFER ---
  "TA",      "buffer",    trans("Bikeways.parquet"),                "1=1", NA,      c(0.25, 0.5, 0.75, 1.5, 3.0)
)
```

## Processing Function

### Helper Function

This function handles the transform, union-by-value optimization, and spatial joins entirely in UTM 12N.

```{r}
#| label: func-spatial-engine

process_layer <- function(conn, target_table, point_table, config) {

  col  <- config$col_name
  path <- config$path
  type <- config$type
  sql_filter <- config$filter_sql

  # Set simplification tolerance based on data type:
  if (type == "isochrone") {
    SIMPLIFICATION_TOLERANCE <- 10.0 # Adjusted to 10m for stabilization
  } else {
    SIMPLIFICATION_TOLERANCE <- 5.0
  }

  if (!fs::file_exists(path)) {
    message("⚠️ Missing file: ", path)
    return(NULL)
  }

  message(glue::glue("🔄 Processing {col} ({type}) for {target_table} with {SIMPLIFICATION_TOLERANCE}m tolerance..."))

  # C. Load Raw Data (UTM + Clean) - CRITICAL FIX
  # Applies axis-fix, validation, and simplification before the join.
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE raw_layer AS
    SELECT
      -- FIX: ST_MakeValid first, then ST_Simplify more aggressively
      ST_Simplify(
        ST_MakeValid(
          ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)
        ),
        {SIMPLIFICATION_TOLERANCE}
      ) as geom
      {if(type=='isochrone') glue::glue(', val') else ''}
    FROM read_parquet('{path}')
    WHERE {sql_filter}
      AND ST_IsValid(geometry)
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_raw_layer ON raw_layer USING RTREE (geom)")

  # D. Add Column if missing (and reset to NULL if exists)
  cols <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {target_table}"))$Field
  if (!col %in% cols) {
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN {col} DOUBLE DEFAULT NULL"))
  } else {
    DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = NULL"))
  }

  # E. Execute Spatial Logic (Binary, Isochrone, Buffer) - (Logic remains the same)
  if (type == "binary") {
    DBI::dbExecute(conn, glue::glue("
      UPDATE {target_table}
      SET {col} = 1
      FROM (
        SELECT DISTINCT tp.rowid
        FROM {point_table} tp, raw_layer r
        WHERE ST_Intersects(tp.geom, r.geom)
      ) match
      WHERE {target_table}.rowid = match.rowid
    "))
    DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = 0 WHERE {col} IS NULL"))

  } else if (type == "isochrone") {
    DBI::dbExecute(conn, glue::glue("
      UPDATE {target_table}
      SET {col} = matched.min_val
      FROM (
        SELECT tp.rowid, MIN(r.val) as min_val
        FROM {point_table} tp, raw_layer r
        WHERE ST_Intersects(tp.geom, r.geom)
        GROUP BY tp.rowid
      ) matched
      WHERE {target_table}.rowid = matched.rowid
    "))

  } else if (type == "buffer") {
    dists_miles <- config$buffer_opts[[1]]
    DBI::dbExecute(conn, "CREATE OR REPLACE TEMP TABLE union_lines AS SELECT ST_MakeValid(ST_Union_Agg(geom)) as geom FROM raw_layer")
    for (miles in sort(dists_miles, decreasing = TRUE)) {
      meters <- miles * 1609.34
      DBI::dbExecute(conn, glue::glue("
        UPDATE {target_table}
        SET {col} = {miles}
        FROM union_lines u, {point_table} tp
        WHERE {target_table}.rowid = tp.rowid
          AND ST_Intersects(tp.geom, ST_Buffer(u.geom, {meters}))
      "))
    }
    DBI::dbExecute(conn, "DROP TABLE union_lines")
  }

  DBI::dbExecute(conn, "DROP TABLE raw_layer;")
}
```

### Spatial Join

We iterate through the config and apply logic to both tables.

```{r}
#| label: run-engine

message("Starting Amenity Batch Processing...")
amenity_config |>
  purrr::pwalk(function(...) {
    row <- list(...)
    process_layer(conn, "targets_parcels", "points_parcels", row)
    process_layer(conn, "targets_h3", "points_h3", row)
  })

message("✅ All Amenity Layers Processed.")
```

# Export

We export the "Wide" tables (ID + all amenity times). We exclude the geometry to keep the files lightweight for the next scoring step.

```{r}
#| label: export

# 1. Parcels
fp_out_parcels <- here::here(dirs$results, "parcels_enriched.parquet")
DBI::dbExecute(conn, glue::glue("COPY (SELECT * FROM targets_parcels) TO '{fp_out_parcels}' (FORMAT PARQUET)"))

# 2. H3
fp_out_h3 <- here::here(dirs$results, "h3_enriched.parquet")
DBI::dbExecute(conn, glue::glue("COPY (SELECT * FROM targets_h3) TO '{fp_out_h3}' (FORMAT PARQUET)"))

message("🚀 Processing Complete!")
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
