---
title: Spatial Operations
subtitle: Enrich Parcels and H3 Grids with Amenity Access
description: Perform spatial joins between target geometries (Parcels, H3) and isochrone layers to calculate access times.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-12"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here"         # https://github.com/r-lib/here/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source     = here::here("_src"), # Source custom functions
  raw        = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed  = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered   = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results    = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base      = here::here(dirs$processed, "base_layers"),
  parcels   = here::here(dirs$processed, "parcels_taz"),
  landuse   = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access    = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-iso-subdirs

# Define Sub-directories for Isochrones
iso_dirs <- list(
  walking   = here::here(dirs$isochrones, "foot-walking"),
  biking    = here::here(dirs$isochrones, "cycling-regular"),
  driving   = here::here(dirs$isochrones, "driving-car")
)

purrr::walk(iso_dirs, ~fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

## Initialize Database

We use DuckDB for all spatial operations to avoid loading millions of geometries into R memory.

```{r}
#| label: init-db

# Create persistent connection for this session
conn <- DBI::dbConnect(duckdb::duckdb())

# Load Spatial Extensions
# Note: Ensure you have the 'spatial' extension installed in DuckDB
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
```

# Part 1: Prepare Targets (The Canvas)

We start by loading the Parcels and H3 grids. We immediately join Administrative and Building attributes so our "Canvas" is fully described before we start painting it with amenities.

## Load Parcels

```{r}
#| label: load-parcels

fp_remm_parcels <- here::here(sub_dirs$parcels, "REMM_BY_parcels.parquet")

# We create a persistent table 'targets_parcels'
# Note: Adjust 'parcel_id' if your REMM data uses 'objectid' or 'parcel_id'
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_parcels AS
  SELECT
    parcel_id,
    ST_GeomFromWKB(geometry) as geom
  FROM read_parquet('{fp_remm_parcels}')
"))

# Spatial Index for performance
DBI::dbExecute(conn, "CREATE INDEX idx_parcels_geom ON targets_parcels USING RTREE (geom)")

message("✅ Parcels loaded into DuckDB")
```

## Load H3 Grid

```{r}
#| label: load-h3

fp_h3_grid <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")

DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_h3 AS
  SELECT
    h3_index,
    ST_GeomFromWKB(geometry) as geom
  FROM read_parquet('{fp_h3_grid}')
"))

# Spatial Index for performance
DBI::dbExecute(conn, "CREATE INDEX idx_h3_geom ON targets_h3 USING RTREE (geom)")

message("✅ H3 Grid loaded into DuckDB")
```

# Part 2: Static Attributes (Admin & Land Use)

## Attach Administrative Data

A helper function to join City/County codes to any target table (Parcel or H3).

```{r}
#| label: func-admin-join

attach_admin_data <- function(conn, target_table) {

  message(glue::glue("🏛️ Attaching Admin Data to {target_table}..."))

  # 1. Municipalities (CommCode)
  fp_muni <- here::here(sub_dirs$base, "UtahMunicipalBoundaries.parquet")

  # Create temp layer in UTM
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_muni AS
    SELECT
      ugrcode as CommCode,
      ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}') as geom
    FROM read_parquet('{fp_muni}')
  "))

  # Spatial Join (Centroid)
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS CommCode VARCHAR;
    UPDATE {target_table}
    SET CommCode = m.CommCode
    FROM layer_muni m
    WHERE ST_Intersects(ST_Centroid({target_table}.geom), m.geom);
  "))

  # 2. Counties (COUNTYNBR)
  fp_county <- here::here(sub_dirs$base, "UtahCountyBoundaries.parquet")

  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_county AS
    SELECT
      countynbr as COUNTYNBR,
      ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}') as geom
    FROM read_parquet('{fp_county}')
  "))

  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS COUNTYNBR VARCHAR;
    UPDATE {target_table}
    SET COUNTYNBR = c.COUNTYNBR
    FROM layer_county c
    WHERE ST_Intersects(ST_Centroid({target_table}.geom), c.geom);
  "))

  DBI::dbExecute(conn, "DROP TABLE layer_muni; DROP TABLE layer_county;")
}
```

```{r}
#| label: run-admin-join
attach_admin_data(conn, "targets_parcels")
attach_admin_data(conn, "targets_h3")
```

## Building Codes (BC)

We join the REMM buildings table. Since `building_type` contains full strings (e.g., "Single Family Res"), we create a mapping table to translate them to abbreviations (e.g., "SF").

```{r}
#| label: building-codes

# 1. Load Buildings Table
fp_buildings <- here::here(sub_dirs$parcels, "REMM_BY_buildings.parquet")

DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TEMP TABLE remm_buildings AS
  SELECT parcel_id, building_type FROM read_parquet('{fp_buildings}')
"))

# 2. Create Mapping Table
bc_mapping <- tibble::tribble(
  ~full_name,                  ~abbr,
  "Agriculture",              "AG",
  "Empty Buildable",          "EM",
  "Open Space Non-Buildable", "OS",
  "Church",                   "CH",
  "Single Family Res",        "SF",
  "Multi Family Res",         "MF",
  "Group Quarters",           "GQ",
  "Governmental",             "GO",
  "Educational",              "ED",
  "Healthcare",               "HE",
  "Retail",                   "RE",
  "Office",                   "OF",
  "Industrial",               "IN",
  "Utilities",                "UT",
  "No Build",                 "NB",
  "Other",                    "OT"
)

# Load mapping to DB
DBI::dbWriteTable(conn, "bc_map", bc_mapping, overwrite = TRUE)

# 3. Join to Parcels
# We join buildings to parcels, then map to abbreviation.
# Note: Parcels might have multiple buildings. We take the 'First' or 'Mode' logic.
# Here we arbitrarily take the first associated building type if duplicates exist.
DBI::dbExecute(conn, "
  ALTER TABLE targets_parcels ADD COLUMN BC VARCHAR;

  UPDATE targets_parcels
  SET BC = mapped.abbr
  FROM (
    SELECT
      b.parcel_id,
      m.abbr
    FROM remm_buildings b
    JOIN bc_map m ON b.building_type = m.full_name
    QUALIFY ROW_NUMBER() OVER (PARTITION BY b.parcel_id) = 1
  ) mapped
  WHERE targets_parcels.parcel_id = mapped.parcel_id
")

DBI::dbExecute(conn, "DROP TABLE remm_buildings; DROP TABLE bc_map;")
```

## Attach Job Access (AA, AT)

We attach the raw job access numbers from the TAZ layer. The normalization (dividing by city max) will happen in Step 4.

```{r}
#| label: func-job-join

attach_job_access <- function(conn, target_table) {

  message(glue::glue("💼 Attaching Job Access to {target_table}..."))

  fp_taz_access <- here::here(dirs$processed, "access", "TAZ_AccessToOpportunities_OY2023.parquet")

  # Load TAZ in UTM
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE taz_access AS
    SELECT
      ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}') as geom,
      job_by_auto,
      job_by_tran
    FROM read_parquet('{fp_taz_access}')
  "))

  # Join
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS AA DOUBLE;
    ALTER TABLE {target_table} ADD COLUMN IF NOT EXISTS AT DOUBLE;

    UPDATE {target_table}
    SET
      AA = t.job_by_auto,
      AT = t.job_by_tran
    FROM taz_access t
    WHERE ST_Intersects(ST_Centroid({target_table}.geom), t.geom);
  "))

  DBI::dbExecute(conn, "DROP TABLE taz_access;")
}
```

```{r}
#| label: run-job-join
attach_job_access(conn, "targets_parcels")
attach_job_access(conn, "targets_h3")
```

# Part 3: Amenity Operations Engine

## Configuration

We define all external layers here. This structure handles Binary (In/Out), Isochrone (Time), and Buffer (Distance) logic.

```{r}
#| label: config-amenities

# Helper for paths
iso_walk  <- function(f) here::here(dirs$isochrones, "foot-walking", f)
iso_drive <- function(f) here::here(dirs$isochrones, "driving-car", f)
land_use  <- function(f) here::here(sub_dirs$landuse, f)
trans     <- function(f) here::here(sub_dirs$transport, f)
amen      <- function(f) here::here(sub_dirs$amenities, f)

amenity_config <- tibble::tribble(
  ~col_name, ~type,       ~path,                                    ~filter_sql,                    ~value_col, ~buffer_opts,

  # --- BINARY LAYERS (0/1) ---
  "OZ",      "binary",    land_use("Utah_Qualified_Opportunity_Zones.parquet"), "1=1",              NA,         NA,
  "AP",      "binary",    amen("WalkAccesstoParks_10Min.parquet"),              "1=1",              NA,         NA,

  # Centers (Split by Type)
  "CM",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Metropolitan Center'", NA, NA,
  "CU",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Urban Center'",        NA, NA,
  "CC",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'City Center'",         NA, NA,
  "CN",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Neighborhood Center'", NA, NA,

  # --- ISOCHRONES (Time Values) ---
  # We extract the raw 'value' (300, 600, etc) here.
  "TT",      "isochrone", iso_walk("Transit_Current.parquet"),      "1=1", "value", NA,
  "TF",      "isochrone", iso_drive("Transit_Current.parquet"),     "1=1", "value", NA,
  "AC",      "isochrone", iso_walk("ChildCare.parquet"),            "1=1", "value", NA,
  "AH",      "isochrone", iso_drive("Healthcare.parquet"),          "1=1", "value", NA,
  "AG",      "isochrone", iso_walk("Grocery.parquet"),              "1=1", "value", NA,
  "AM",      "isochrone", iso_walk("Community_Centers.parquet"),    "1=1", "value", NA,
  "AE_K12",  "isochrone", iso_walk("Schools_PreK12.parquet"),       "1=1", "value", NA,
  "AE_High", "isochrone", iso_drive("Schools_HigherEd.parquet"),    "1=1", "value", NA,

  # --- BUFFER LAYERS (Distance in Miles) ---
  # We will convert these miles to meters inside the function for UTM buffering
  "TA",      "buffer",    trans("Bikeways.parquet"),                "1=1", NA,      c(0.25, 0.5, 1.0) # miles
)
```

## Processing Function

### Helper Function

This function handles the transform, union-by-value optimization, and spatial joins entirely in UTM 12N.

```{r}
#| label: func-spatial-engine

process_layer <- function(conn, target_table, config) {

  col  <- config$col_name
  path <- config$path
  type <- config$type
  sql_filter <- config$filter_sql

  if (!fs::file_exists(path)) {
    message("⚠️ Missing file: ", path)
    return(NULL)
  }

  message(glue::glue("🔄 Processing {col} ({type}) for {target_table}..."))

  # 1. Load Raw Data & Transform to UTM immediately
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE raw_layer AS
    SELECT *, ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}') as geom
    FROM read_parquet('{path}')
    WHERE {sql_filter}
  "))

  # 2. Add Column if missing
  cols <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {target_table}"))$Field
  if (!col %in% cols) {
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN {col} DOUBLE DEFAULT NULL"))
  }

  # 3. Execution Strategy

  if (type == "binary") {
    # OPTIMIZATION: Union Aggregation to create a single complex geometry to test against
    DBI::dbExecute(conn, glue::glue("
      UPDATE {target_table}
      SET {col} = 1
      FROM (SELECT ST_Union_Agg(geom) as geom FROM raw_layer) poly
      WHERE ST_Intersects({target_table}.geom, poly.geom)
    "))
    DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = 0 WHERE {col} IS NULL"))

  } else if (type == "isochrone") {
    # OPTIMIZATION: Group by value (300, 600) and Union.
    # Reduces N polygons to just K polygons (where K = number of time bands).
    val_col <- config$value_col

    DBI::dbExecute(conn, glue::glue("
      CREATE OR REPLACE TEMP TABLE optimized_iso AS
      SELECT {val_col} as val, ST_Union_Agg(geom) as geom
      FROM raw_layer
      GROUP BY {val_col}
    "))

    # Update with MIN value
    DBI::dbExecute(conn, glue::glue("
      UPDATE {target_table}
      SET {col} = matched.min_val
      FROM (
        SELECT t.rowid as tid, MIN(iso.val) as min_val
        FROM {target_table} t, optimized_iso iso
        WHERE ST_Intersects(t.geom, iso.geom)
        GROUP BY t.rowid
      ) matched
      WHERE {target_table}.rowid = matched.tid
    "))
    DBI::dbExecute(conn, "DROP TABLE optimized_iso")

  } else if (type == "buffer") {
    # BUFFER: Create buffers in Meters (UTM)
    dists_miles <- config$buffer_opts[[1]]

    # Pre-aggregate lines for speed (reduces buffer operations)
    DBI::dbExecute(conn, "CREATE OR REPLACE TEMP TABLE union_lines AS SELECT ST_Union_Agg(geom) as geom FROM raw_layer")

    # Iterate buffers (Largest to Smallest? Or just update all matching?)
    # Since we want the *smallest* distance (closest access),
    # we can iterate decreasing distance. If a parcel hits 1.0, it gets 1.0.
    # If it also hits 0.25, it gets overwritten with 0.25.

    for (miles in sort(dists_miles, decreasing = TRUE)) {
      meters <- miles * 1609.34

      DBI::dbExecute(conn, glue::glue("
        UPDATE {target_table}
        SET {col} = {miles}
        FROM union_lines u
        WHERE ST_Intersects(
          {target_table}.geom,
          ST_Buffer(u.geom, {meters})
        )
      "))
    }
    DBI::dbExecute(conn, "DROP TABLE union_lines")
  }

  DBI::dbExecute(conn, "DROP TABLE raw_layer")
}
```

### Spatial Join

We iterate through the config and apply logic to both tables.

```{r}
#| label: run-engine

# 1. Run Standard Layers
amenity_config |>
  purrr::pwalk(function(...) {
    row <- list(...)

    # Run for Parcels
    process_layer(conn, "targets_parcels", row)

    # Run for H3 (excluding parcel-specific logic if any, but amenities apply to all)
    process_layer(conn, "targets_h3", row)
  })

# 2. Post-Process Education (AE) for Parcels
DBI::dbExecute(conn, "
  ALTER TABLE targets_parcels ADD COLUMN IF NOT EXISTS AE DOUBLE;
  UPDATE targets_parcels SET AE = LEAST(AE_K12, AE_High);
  ALTER TABLE targets_parcels DROP COLUMN AE_K12;
  ALTER TABLE targets_parcels DROP COLUMN AE_High;
")

# 3. Post-Process Education (AE) for H3
DBI::dbExecute(conn, "
  ALTER TABLE targets_h3 ADD COLUMN IF NOT EXISTS AE DOUBLE;
  UPDATE targets_h3 SET AE = LEAST(AE_K12, AE_High);
  ALTER TABLE targets_h3 DROP COLUMN AE_K12;
  ALTER TABLE targets_h3 DROP COLUMN AE_High;
")
```

# Export

We export the "Wide" tables (ID + all amenity times). We exclude the geometry to keep the files lightweight for the next scoring step.

```{r}
#| label: export

# 1. Parcels
fp_out_parcels <- here::here(dirs$results, "parcels_enriched.parquet")
DBI::dbExecute(conn, glue::glue("
  COPY (SELECT * FROM targets_parcels)
  TO '{fp_out_parcels}' (FORMAT PARQUET)
"))

# 2. H3
fp_out_h3 <- here::here(dirs$results, "h3_enriched.parquet")
DBI::dbExecute(conn, glue::glue("
  COPY (SELECT * FROM targets_h3)
  TO '{fp_out_h3}' (FORMAT PARQUET)
"))

message("🚀 Processing Complete!")
message("Output Parcels: ", fp_out_parcels)
message("Output H3:      ", fp_out_h3)
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
