---
title: Spatial Operations
subtitle: Lorem Ipsum
description: Conduct Spatial Operations to calculate the final results.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

---
title: Process Data Layers
subtitle: Lorem Ipsum
description: Process the data to prepare for spatial operations (buffer, isochrones, etc.)
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here"         # https://github.com/r-lib/here/
    ),
    dependencies = TRUE
  )
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(sf)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Database Setup

```{r}
#| label: setup-db-init

# Initialize In-memory database
conn <- DBI::dbConnect(duckdb::duckdb())

# install and load spatial extension
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extension()")
```

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source = here::here("_src"), # Source custom functions
  raw = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results = here::here("_output") # Final Output: Analysis Results
)
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base   = here::here(dirs$processed, "base_layers"),
  parcels = here::here(dirs$processed, "parcels_taz"),
  landuse = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_PROJ <- "EPSG:3566" # NAD83 / Utah Central (Projected - ftUS)
CRS_VIEW <- "EPSG:3857" # WGS 84 / Pseudo-Mercator (Projected - Meters)
CRS_WGS  <- "EPSG:4326" # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "download_feature_layer.R")) # Download ArcGIS Feature Layers
source(here::here(dirs$source, "process_local_layer.R")) # Load local SHP, GDB, GeoJSON Layers
source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```
