---
title: Spatial Operations
subtitle: Enrich Parcels and H3 Grids with Amenity Access
description: Perform spatial joins between target geometries (Parcels, H3) and isochrone layers to calculate access times.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-12"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here",        # https://github.com/r-lib/here/
    "fs",          # https://github.com/r-lib/fs/
    "zip"          # https://github.com/r-lib/zip
    ),
    dependencies = TRUE
  )
```

OR,

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak("renv", dependencies = TRUE) # https://github.com/rstudio/renv/

renv::restore()
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source     = here::here("_src"), # Source custom functions
  raw        = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed  = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered   = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results    = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base      = here::here(dirs$processed, "base_layers"),
  parcels   = here::here(dirs$processed, "parcels_taz"),
  landuse   = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access    = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-iso-subdirs

# Define Sub-directories for Isochrones
iso_dirs <- list(
  walking   = here::here(dirs$isochrones, "foot-walking"),
  biking    = here::here(dirs$isochrones, "cycling-regular"),
  driving   = here::here(dirs$isochrones, "driving-car")
)

purrr::walk(iso_dirs, ~fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

## Initialize Database

We use DuckDB for all spatial operations to avoid loading millions of geometries into R memory.

```{r}
#| label: init-db

# Create persistent connection for this session
conn <- DBI::dbConnect(duckdb::duckdb())

# Load Spatial Extensions
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
```

# Part 1: Prepare Targets (The Canvas)

We start by loading the Parcels and H3 grids. We immediately join Administrative and Building attributes so our "Canvas" is fully described before we start painting it with amenities.

Check Desired format here: https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/HousingSuitability_Centers2025_gdb/FeatureServer/0

## Load H3 Grid

We load H3 first to get the resolution, then use that resolution to load Parcels with the correct H3 index pre-calculated.

```{r}
#| label: load-h3

fp_h3_grid <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")

# 1. Load Polygons (UTM)
#    Note: H3 WKB is usually WGS84, so we must transform to UTM.
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_h3 AS
  SELECT
    CAST(h3_index AS UBIGINT) as h3_index, -- Strict type safety
    ST_Transform(ST_GeomFromWKB(geometry), '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
  FROM read_parquet('{fp_h3_grid}')
"))
DBI::dbExecute(conn, "CREATE INDEX idx_h3_geom ON targets_h3 USING RTREE (geom)")

# 2. OPTIMIZATION: Create & Index Points
message("📍 Pre-calculating H3 Points (PointOnSurface)...")
DBI::dbExecute(conn, "
  CREATE OR REPLACE TABLE points_h3 AS
  SELECT rowid, h3_index, ST_Centroid(geom) as geom
  FROM targets_h3
")
DBI::dbExecute(conn, "CREATE INDEX idx_h3_utm ON points_h3 USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_h3_id ON points_h3 (h3_index)")

message("✅ H3 Grid loaded & indexed")
```

```{r}
# -----------------------------------------------------------------------------
# 2. GET RESOLUTION
# -----------------------------------------------------------------------------
# We need this 'res' variable for the next steps
res <- DBI::dbGetQuery(conn, "SELECT h3_get_resolution(h3_index) as res FROM targets_h3 LIMIT 1")$res
message(glue::glue("ℹ️ Detected Grid Resolution: {res}"))
```

## Load Parcels

```{r}
#| label: load-parcels

fp_remm_parcels <- here::here(sub_dirs$parcels, "REMM_BY_parcels.parquet")

# 1. Load Polygons (UTM)
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE targets_parcels AS
  SELECT
    parcel_id,

    -- 1. Calculate H3 Index DIRECTLY from raw WGS84 Centroid
    CAST(h3_latlng_to_cell(
      ST_Y(ST_Centroid(geometry)),
      ST_X(ST_Centroid(geometry)),
      {res}
    ) AS UBIGINT) as h3_index,

    -- 2. Transform to UTM for storage/buffer ops
    ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)) as geom
  FROM read_parquet('{fp_remm_parcels}')
  WHERE ST_IsValid(geometry)
"))

# 2. OPTIMIZATION: Create & Index Points (Point-On-Surface)
#    ST_PointOnSurface guarantees the point is INSIDE the polygon (unlike Centroid).
#    This handles U-shaped parcels correctly.
DBI::dbExecute(conn, "
  CREATE OR REPLACE TABLE points_parcels AS
  SELECT rowid, h3_index, ST_PointOnSurface(geom) as geom
  FROM targets_parcels
")
DBI::dbExecute(conn, "CREATE INDEX idx_pp_utm ON points_parcels USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_pp_h3 ON points_parcels (h3_index)")

message("✅ Parcels loaded & indexed")
```

# Part 2: Static Attributes (Admin & Land Use)

## Attach Administrative Data

A helper function to join City/County codes to any target table (Parcel or H3).

```{r}
#| label: func-admin-join

attach_admin_data <- function(conn, target_table, point_table) {

  message(glue::glue("🏛️ Attaching Admin Data to {target_table}..."))
  fp_municipal_bound <- here::here(sub_dirs$base, "UtahMunicipalBoundaries.parquet")
  fp_county_bound    <- here::here(sub_dirs$base, "UtahCountyBoundaries.parquet")

  # 1. Load Admin Layers (Fixed Geometry)
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_muni AS
    SELECT
      ugrcode as CommCode,
      ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_municipal_bound}')
  "))

  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_county AS
    SELECT
      countynbr as COUNTYNBR,
      ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_county_bound}')
  "))

  # 2. Add Columns
  for (col in c("CommCode", "COUNTYNBR")) {
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} DROP COLUMN IF EXISTS {col}"))
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN {col} VARCHAR"))
  }

  # 3. Update City (Spatial Join on Point Table -> Update Target by RowID)
  DBI::dbExecute(conn, glue::glue("
    UPDATE {target_table}
    SET CommCode = match.CommCode
    FROM (
        SELECT pt.rowid, m.CommCode
        FROM {point_table} pt, layer_muni m
        WHERE ST_Intersects(pt.geom, m.geom)
    ) match
    WHERE {target_table}.rowid = match.rowid
  "))

  # 4. Update County (Spatial Join on Point Table -> Update Target by RowID)
  DBI::dbExecute(conn, glue::glue("
    UPDATE {target_table}
    SET COUNTYNBR = match.COUNTYNBR
    FROM (
        SELECT pt.rowid, c.COUNTYNBR
        FROM {point_table} pt, layer_county c
        WHERE ST_Intersects(pt.geom, c.geom)
    ) match
    WHERE {target_table}.rowid = match.rowid
  "))

  DBI::dbExecute(conn, "DROP TABLE layer_muni; DROP TABLE layer_county;")
}
```

```{r}
#| label: run-admin-join

# Pass the pre-calculated point tables we created in Part 1
attach_admin_data(conn, "targets_parcels", "points_parcels")
attach_admin_data(conn, "targets_h3", "points_h3")
```

## Building Codes (BC)

We join the REMM buildings table. Since `building_type` contains full strings (e.g., "Single Family Res"), we create a mapping table to translate them to abbreviations (e.g., "SF").

```{r}
#| label: building-codes

# -----------------------------------------------------------------------------
# 1. Load Buildings & Store as UTM (Projected)
# -----------------------------------------------------------------------------
message("🏗️ Loading Building Centroids (Spatial)...")

fp_remm_buildings <- here::here(sub_dirs$parcels, "REMM_BY_buildings.parquet")

# Create Temp Table with BOTH UTM Geom and H3 Index
# Load -> Centroid -> Transform (UTM) -> Index
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TEMP TABLE points_buildings AS
  SELECT
    parcel_id,
    building_type,
    building_sqft,

    -- 1. H3 Index (Integer Match)
    CAST(h3_latlng_to_cell(
      ST_Y(ST_Centroid(geometry)),
      ST_X(ST_Centroid(geometry)),
      {res}
    ) AS UBIGINT) as h3_index,

    -- 2. UTM Geometry (Spatial Match)
    ST_Transform(ST_Centroid(geometry), '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom

  FROM read_parquet('{fp_remm_buildings}')
  WHERE ST_IsValid(geometry)
"))

# Index both columns for speed
DBI::dbExecute(conn, "CREATE INDEX idx_b_utm ON points_buildings USING RTREE (geom)")
DBI::dbExecute(conn, "CREATE INDEX idx_b_h3 ON points_buildings (h3_index)")

# Create Mapping Table (Full Name -> Abbreviation)
bc_mapping <- tibble::tribble(
  ~full_name,                  ~abbr,
  "Agriculture",              "AG",
  "Church",                   "CH",
  "Commons",                  "OS",
  "Educational",              "ED",
  "Empty - Buildable",        "EM",
  "Empty Buildable",          "EM",
  "Governmental",             "GO",
  "Group Quarters",           "GQ",
  "Healthcare",               "HE",
  "Industrial",               "IN",
  "Mixed Use",                "MU",
  "Multi Family Res",         "MF",
  "No Build",                 "NB",
  "Office",                   "OF",
  "Open Space Non-Buildable", "OS",
  "Other",                    "OT",
  "Retail",                   "RE",
  "Single Family Res",        "SF",
  "Utilities",                "UT"
  # Note: NA values are handled in the join below
)
DBI::dbWriteTable(conn, "bc_map", bc_mapping, overwrite = TRUE)

# -----------------------------------------------------------------------------
# 2. PARCELS: Attribute Join (Parcel ID Match)
# -----------------------------------------------------------------------------
message("🏗️ Attaching Building Codes to Parcels (Attribute Match)...")

# Join Abbreviation to Parcels
# Note: QUALIFY ROW_NUMBER() handles cases where a parcel has multiple buildings
# Note: We prioritize the LARGEST building on the parcel using building_sqft
DBI::dbExecute(conn, "
  ALTER TABLE targets_parcels ADD COLUMN BC VARCHAR;

  UPDATE targets_parcels
  SET BC = COALESCE(mapped.abbr, 'OT') -- Handle NA/Unknown as 'OT'
  FROM (
    SELECT
      b.parcel_id,
      m.abbr
    FROM points_buildings b
    JOIN bc_map m ON b.building_type = m.full_name

    -- Order by square footage descending to pick the main building
    QUALIFY ROW_NUMBER() OVER (
      PARTITION BY b.parcel_id
      ORDER BY b.building_sqft DESC
    ) = 1
  ) mapped
  WHERE targets_parcels.parcel_id = mapped.parcel_id
")

# -----------------------------------------------------------------------------
# 3. H3: Hybrid Approach
#    Pass A: Building Aggregation (Handles small parcels / mixed use)
#    Pass B: Spatial Backfill (Handles large parcels / empty lots)
# -----------------------------------------------------------------------------
message("🏗️ Aggregating Building Codes to H3 (Index Match)...")

# --- Pass A: Aggregation via Index Match (Existing Logic) ---
DBI::dbExecute(conn, glue::glue("
  ALTER TABLE targets_h3 ADD COLUMN IF NOT EXISTS BC VARCHAR;

  UPDATE targets_h3
  SET BC = agg.mode_bc
  FROM (
    SELECT
      b.h3_index,
      MODE(COALESCE(m.abbr, 'OT')) as mode_bc
    FROM points_buildings b
    LEFT JOIN bc_map m ON b.building_type = m.full_name
    GROUP BY b.h3_index
  ) agg
  WHERE targets_h3.h3_index = agg.h3_index
"))

# --- Pass B: Spatial Backfill (Gap Filling) ---
# Check: If BC is still NULL, sample the Parcel Polygon at the H3 Center.
# This fixes large parcels where the H3 cell is far from the building centroid.

message("  ↳ Filling gaps using Parcel Polygons...")

# 1. Ensure Polygon Index Exists (Critical for spatial join speed)
DBI::dbExecute(conn, "CREATE INDEX IF NOT EXISTS idx_tp_geom ON targets_parcels USING RTREE (geom)")

# 2. Update NULLs only
DBI::dbExecute(conn, "
  UPDATE targets_h3
  SET BC = matched.BC
  FROM (
    SELECT
      ph.h3_index,
      p.BC
    FROM points_h3 ph
    -- Spatial Join: H3 Center inside Parcel Polygon
    JOIN targets_parcels p ON ST_Intersects(ph.geom, p.geom)
    WHERE p.BC IS NOT NULL
  ) matched
  WHERE targets_h3.h3_index = matched.h3_index
    AND targets_h3.BC IS NULL -- Only touch rows that failed Pass A
")

# --- Pass C: Spatial Voting (KNN Mode) ---
# Logic: If a cell is empty, look at its 6 immediate neighbors.
# If 4 or more neighbors have data, we assume this is a "glitch" or hole -> FILL IT.
# If 3 or fewer neighbors have data, we assume this is a coastline/edge -> LEAVE IT.
message("  ↳ Pass C: Filling holes using Grid Adjacency (Threshold >= 4)...")

DBI::dbExecute(conn, "
  UPDATE targets_h3
  SET BC = fill.mode_bc
  FROM (
    SELECT
      t.h3_index,
      MODE(n.BC) as mode_bc
    FROM targets_h3 t
    -- 1. Get the 6 immediate neighbors (k=1) for the empty cell
    CROSS JOIN LATERAL (
      SELECT unnest(h3_grid_disk(t.h3_index, 1)) as neighbor_h3
    ) grid
    -- 2. Join back to the main table to check the neighbor's value
    JOIN targets_h3 n ON grid.neighbor_h3 = n.h3_index

    WHERE t.BC IS NULL          -- Target must be empty
      AND n.BC IS NOT NULL      -- Neighbor must have data
      AND t.h3_index != n.h3_index -- Exclude the cell itself

    GROUP BY t.h3_index
    -- 3. THE RULE: Only fill if surrounded by at least 4 valid cells
    HAVING COUNT(*) >= 4
  ) fill
  WHERE targets_h3.h3_index = fill.h3_index
")

# --- Pass D: Final Cleanup (Catch-All) ---
# Logic: If it is STILL null after Pass C, it means it is either:
# 1. Genuinely empty (surrounded by < 4 valid neighbors)
# 2. A remote area with no data
# We force these to 'OT' (Other).
message("  ↳ Pass D: Setting remaining NULLs to 'OT'...")

DBI::dbExecute(conn, "
  UPDATE targets_h3
  SET BC = 'OT'
  WHERE BC IS NULL
")

# Cleanup
DBI::dbExecute(conn, "DROP TABLE points_buildings; DROP TABLE bc_map;")
```

```{r}
#| label: check-missing-b3
#| eval: false

# Check H3 for missing BC values
DBI::dbGetQuery(conn, "
  SELECT
    COUNT(*) as total,
    COUNT(BC) as filled,
    COUNT(*) - COUNT(BC) as missing,
    ROUND(100.0 * (COUNT(*) - COUNT(BC)) / COUNT(*), 2) as missing_pct
  FROM targets_h3
")

# Check Parcels (for comparison)
parcel_stats <- DBI::dbGetQuery(conn, "
  SELECT
    COUNT(*) as total,
    COUNT(BC) as filled,
    COUNT(*) - COUNT(BC) as missing,
    ROUND(100.0 * (COUNT(*) - COUNT(BC)) / COUNT(*), 2) as missing_pct
  FROM targets_parcels
")
```

## Attach Job Access (AA, AT)

We attach the raw job access numbers from the TAZ layer. The normalization (dividing by city max) will happen in Step 4.

```{r}
#| label: func-job-join

fp_taz_access <- here::here(dirs$processed, "access", "TAZ_AccessToOpportunities_OY2023.parquet")

attach_job_access <- function(conn, target_table, point_table) {

  message(glue::glue("💼 Attaching Job Access to {target_table}..."))

  # 1. Load TAZ & SAFE Transform
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE taz_access AS
    SELECT
      ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)) as geom,
      job_by_auto,
      job_by_tran
    FROM read_parquet('{fp_taz_access}')
    WHERE ST_IsValid(geometry)
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_taz_geom ON taz_access USING RTREE (geom)")

  # 2. RESET COLUMNS (Drop if exists)
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} DROP COLUMN IF EXISTS \"AA\""))
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} DROP COLUMN IF EXISTS \"AT\""))

  # 3. Add Columns
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN \"AA\" DOUBLE"))
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN \"AT\" DOUBLE"))

  # 4. Spatial Join
  DBI::dbExecute(conn, glue::glue("
    UPDATE {target_table}
    SET \"AA\" = match.job_by_auto, \"AT\" = match.job_by_tran
    FROM (
      SELECT tp.rowid, t.job_by_auto, t.job_by_tran
      FROM {point_table} tp, taz_access t
      WHERE ST_Intersects(tp.geom, t.geom)
    ) match
    WHERE {target_table}.rowid = match.rowid;
  "))

  DBI::dbExecute(conn, "DROP TABLE taz_access;")
}
```

```{r}
#| label: run-job-join

# Run with pre-calculated points
attach_job_access(conn, "targets_parcels", "points_parcels")
attach_job_access(conn, "targets_h3", "points_h3")
```

# Part 3: Amenity Operations Engine

## Configuration

We define all external layers here. This structure handles Binary (In/Out), Isochrone (Time), and Buffer (Distance) logic.

```{r}
#| label: config-amenities

# Helpers
iso_walk  <- function(f) here::here(dirs$isochrones, "foot-walking", f)
iso_bike  <- function(f) here::here(dirs$isochrones, "cycling-regular", f)
iso_drive <- function(f) here::here(dirs$isochrones, "driving-car", f)
land_use  <- function(f) here::here(sub_dirs$landuse, f)
trans     <- function(f) here::here(sub_dirs$transport, f)
amen      <- function(f) here::here(sub_dirs$amenities, f)

amenity_config <- tibble::tribble(
  ~col_name, ~type,       ~path,                                    ~filter_sql,               ~value_col, ~buffer_opts,

  # --- BINARY ---
  "OZ",      "binary",    land_use("Utah_Qualified_Opportunity_Zones.parquet"), "1=1",              NA,         NA,
  "AP",      "binary",    amen("WalkAccesstoParks_10Min.parquet"),              "1=1",              NA,         NA,
  "CM",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Metropolitan Center'", NA, NA,
  "CU",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Urban Center'",        NA, NA,
  "CC",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'City Center'",         NA, NA,
  "CN",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Neighborhood Center'", NA, NA,

  # --- ISOCHRONES ---
  "TT_Local", "isochrone", iso_walk("Transit_Local.parquet"),       "1=1", "value", NA,
  "TT_Rapid", "isochrone", iso_walk("Transit_Rapid.parquet"),       "1=1", "value", NA,
  "TT_CRT",   "isochrone", iso_bike("Transit_CRT.parquet"),         "1=1", "value", NA,
  "TF",      "isochrone", iso_drive("Freeways.parquet"),            "1=1", "value", NA,
  "AC",      "isochrone", iso_walk("ChildCare.parquet"),            "1=1", "value", NA,
  "AH",      "isochrone", iso_drive("Healthcare.parquet"),          "1=1", "value", NA,
  "AG",      "isochrone", iso_walk("Grocery.parquet"),              "1=1", "value", NA,
  "AM",      "isochrone", iso_walk("Community_Centers.parquet"),    "1=1", "value", NA,
  "AE_K12",  "isochrone", iso_walk("Schools_PreK12.parquet"),       "1=1", "value", NA,
  "AE_High", "isochrone", iso_bike("Schools_HigherEd.parquet"),     "1=1", "value", NA,

  # --- BUFFER ---
  "TA_class1", "buffer", trans("Bikeways.parquet"),
  "facility1 LIKE '%(1A)%' OR facility1 LIKE '%(1B)%'",
  NA, c(0.25, 0.5, 0.75, 1.5, 3.0),

  "TA_class2", "buffer", trans("Bikeways.parquet"),
  "facility1 LIKE '%(2A)%' OR facility1 LIKE '%(2B)%' OR facility1 LIKE '%2C%'",
  NA, c(0.25, 0.5, 0.75, 1.5, 3.0),

  "TA_trails", "buffer", trans("Bikeways.parquet"),
  "facility1 LIKE '%Trail%'",
  NA, c(0.25, 0.5, 0.75, 1.5, 3.0)
)
```

## Processing Function

### Helper Function

This function handles the transform, union-by-value optimization, and spatial joins entirely in UTM 12N.

```{r}
#| label: func-spatial-engine
process_layers <- function(conn, point_table, config) {

  col  <- config$col_name
  path <- config$path
  type <- config$type
  sql_filter <- config$filter_sql
  val_col <- config$value_col

  # Simplification Tolerance
  SIMPLIFICATION_TOLERANCE <- 1.0

  if (!fs::file_exists(path)) {
    message("⚠️ Missing file: ", path)
    return(NULL)
  }

  message(glue::glue("🔄 Processing {col} ({type}) on {point_table}..."))

  # A. Load Raw Data (UTM + Clean)
  #    We create the index here, which is CRITICAL for ST_DWithin to be fast.
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE raw_layer AS
    SELECT
      ST_Simplify(
        ST_MakeValid(
          ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)
        ),
        {SIMPLIFICATION_TOLERANCE}
      ) as geom
      {if(type=='isochrone') glue::glue(', {val_col} as val') else ''}
    FROM read_parquet('{path}')
    WHERE {sql_filter}
      AND ST_IsValid(geometry)
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_raw_layer ON raw_layer USING RTREE (geom)")

  # B. Reset Column (Clean Slate)
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {point_table} DROP COLUMN IF EXISTS {col}"))
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {point_table} ADD COLUMN {col} DOUBLE DEFAULT NULL"))

  # C. Execute Spatial Logic
  if (type == "binary") {
    # BINARY: Simple Intersection (Uses Index)
    DBI::dbExecute(conn, glue::glue("
      UPDATE {point_table}
      SET {col} = 1
      FROM raw_layer r
      WHERE ST_Intersects({point_table}.geom, r.geom)
    "))
    DBI::dbExecute(conn, glue::glue("UPDATE {point_table} SET {col} = 0 WHERE {col} IS NULL"))

  } else if (type == "isochrone") {
    # ISOCHRONE: Union & Iterate (Optimized)

    DBI::dbExecute(conn, "
      CREATE OR REPLACE TEMP TABLE grouped_iso AS
      SELECT val, ST_MakeValid(ST_Union_Agg(geom)) as geom
      FROM raw_layer
      GROUP BY val
    ")

    vals <- DBI::dbGetQuery(conn, "SELECT DISTINCT val FROM grouped_iso ORDER BY val ASC")$val

    for (v in vals) {
      DBI::dbExecute(conn, glue::glue("
        UPDATE {point_table}
        SET {col} = {v}
        FROM grouped_iso g
        WHERE {point_table}.{col} IS NULL
          AND g.val = {v}
          AND ST_Intersects({point_table}.geom, g.geom)
      "))
    }
    DBI::dbExecute(conn, "DROP TABLE grouped_iso")

  } else if (type == "buffer") {
    # BUFFER: The Ultra-Fast Method (ST_DWithin)
    # Sorted Distance Loop (Fastest for Nearest Bin)
    dists <- dists <- sort(unlist(config$buffer_opts), decreasing = FALSE)

    for (miles in dists) {
      meters <- miles * 1609.34

      message(glue::glue("   Checking buffer: {miles} miles ({round(meters)}m)"))

      # ST_DWithin uses the RTREE index on raw_layer.
      # The 'WHERE {col} IS NULL' ensures we don't re-calculate points
      # that already matched a closer distance.
      DBI::dbExecute(conn, glue::glue("
        UPDATE {point_table}
        SET {col} = {miles}
        WHERE {col} IS NULL
          AND EXISTS (
             SELECT 1 FROM raw_layer r
             WHERE ST_DWithin({point_table}.geom, r.geom, {meters})
          )
      "))
    }
  }

  DBI::dbExecute(conn, "DROP TABLE raw_layer;")
}
```

### Spatial Join

We iterate through the config and apply logic to both tables.

```{r}
#| label: run-engine

message("🚀 Starting Staged Processing on Points...")

amenity_config |>
  purrr::pwalk(function(...) {
    process_layers(conn, "points_parcels", list(...))
    process_layers(conn, "points_h3", list(...))
  })

message("✅ Points enriched successfully.")
```

```{r}
#| label: transfer-attributes

message("📦 Transferring attributes from Points to Polygons...")

transfer_columns <- function(conn, target_poly, source_point, config) {

  config |>
    purrr::pwalk(function(...) {
      row <- list(...)
      col <- row$col_name

      message(glue::glue("  ↳ Transferring {col}..."))

      # 1. RESET COLUMN on Target (Drop if exists)
      DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_poly} DROP COLUMN IF EXISTS {col}"))
      DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_poly} ADD COLUMN {col} DOUBLE"))

      # 2. Update Target from Source using RowID
      DBI::dbExecute(conn, glue::glue("
        UPDATE {target_poly}
        SET {col} = source.{col}
        FROM {source_point} source
        WHERE {target_poly}.rowid = source.rowid
      "))
    })
}

# Run Transfer
transfer_columns(conn, "targets_parcels", "points_parcels", amenity_config)
transfer_columns(conn, "targets_h3", "points_h3", amenity_config)

message("🏁 All Data Transferred. Processing Complete.")
```

# Export

We export the "Wide" tables (ID + all amenity times). We exclude the geometry to keep the files lightweight for the next scoring step.

```{r}
#| label: export

# 1. Parcels
fp_out_parcels <- here::here(dirs$results, "parcels_enriched.parquet")

# Transform geom from UTM to WGS84 before export
DBI::dbExecute(conn, glue::glue("
  COPY (
    SELECT * EXCLUDE(geom), ST_Transform(geom, '{CRS_UTM}', '{CRS_WGS}', always_xy := true) AS geom
    FROM targets_parcels
  ) TO '{fp_out_parcels}' (FORMAT PARQUET)
"))

# 2. H3
fp_out_h3 <- here::here(dirs$results, "h3_enriched.parquet")

# Transform geom from UTM to WGS84 before export
DBI::dbExecute(conn, glue::glue("
  COPY (
    SELECT * EXCLUDE(geom), ST_Transform(geom, '{CRS_UTM}', '{CRS_WGS}', always_xy := true) AS geom
    FROM targets_h3
  ) TO '{fp_out_h3}' (FORMAT PARQUET)
"))

message("🚀 Processing Complete!")
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
