---
title: Spatial Operations
subtitle: Enrich Parcels and H3 Grids with Amenity Access
description: Perform spatial joins between target geometries (Parcels, H3) and isochrone layers to calculate access times.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-12"
---

# Environment Setup

## Install and Load Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here",        # https://github.com/r-lib/here/
    "fs",          # https://github.com/r-lib/fs/
    "zip"          # https://github.com/r-lib/zip
    ),
    dependencies = TRUE
  )
```

OR,

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak("renv", dependencies = TRUE) # https://github.com/rstudio/renv/

renv::restore()
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(sf)
library(openrouteservice)
library(DBI)
library(duckdb)
library(duckplyr)
library(duckspatial)
library(arrow)
library(geoarrow)
library(janitor)
library(mapgl)
library(here)
```

## Project Configurations

### Path Configuration

```{r}
#| label: setup-dirs

# Setup Paths & Params
dirs <- list(
  source     = here::here("_src"), # Source custom functions
  raw        = here::here("_data/raw"), # Input: Shapefiles, GDBs, CSVs
  processed  = here::here("_data/processed"), # Intermediate Output: Standardized Parquets
  buffered   = here::here("_data/buffered"), # Intermediate Output: Buffered layers
  isochrones = here::here("_data/isochrones"), # Intermediate Output: ORS Isochrones
  results    = here::here("_output") # Final Output: Analysis Results
)

purrr::walk(dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-subdirs

# Define Sub-directories (To match your existing structure)
sub_dirs <- list(
  base      = here::here(dirs$processed, "base_layers"),
  parcels   = here::here(dirs$processed, "parcels_taz"),
  landuse   = here::here(dirs$processed, "landuse_zones"),
  transport = here::here(dirs$processed, "transportation"),
  access    = here::here(dirs$processed, "access"),
  amenities = here::here(dirs$processed, "amenities")
)

purrr::walk(sub_dirs, ~fs::dir_create(., recurse = TRUE))
```

```{r}
#| label: setup-iso-subdirs

# Define Sub-directories for Isochrones
iso_dirs <- list(
  walking   = here::here(dirs$isochrones, "foot-walking"),
  biking    = here::here(dirs$isochrones, "cycling-regular"),
  driving   = here::here(dirs$isochrones, "driving-car")
)

purrr::walk(iso_dirs, ~fs::dir_create(., recurse = TRUE))
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Helper Functions

```{r}
#| label: load-helpers

source(here::here(dirs$source, "preview_layer.R")) # Preview (geo)parquet files using mapgl
```

## Initialize Database

We use DuckDB for all spatial operations to avoid loading millions of geometries into R memory.

```{r}
#| label: init-db

# Create persistent connection for this session
conn <- DBI::dbConnect(duckdb::duckdb())

# Load Spatial Extensions
DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
```

```{r}
#| label: config-db

# 1. Config DuckDB Resource Management
# RAM: 64GB Total -> 40GB for DuckDB, ~20GB for Windows/R overhead (Safe)
DBI::dbExecute(conn, "SET memory_limit = '40GB';")

# Threads: You have 20 logical cores.
# Using 16 leaves 4 for the OS/Background, maximizing spatial join speed.
DBI::dbExecute(conn, "SET threads = 16;")

# Temp Dir: Crucial for preventing OOM crashes.
# Ensure this folder exists on your fastest drive (NVMe SSD).
DBI::dbExecute(conn, "SET temp_directory = 'C:/duckdb_tmp';")
```

# Part 1: Prepare Targets (The Canvas)

We start by loading the Parcels and H3 grids. We immediately join Administrative and Building attributes so our "Canvas" is fully described before we start painting it with amenities.

Check Desired format here: https://services1.arcgis.com/taguadKoI1XFwivx/arcgis/rest/services/HousingSuitability_Centers2025_gdb/FeatureServer/0

```{r}
#| label: target-paths

# Inputs
fp_h3_grid        <- here::here(sub_dirs$base, "Analysis_Boundary_H3_Grid.parquet")
fp_parcels_pts    <- here::here(sub_dirs$parcels, "REMM_BY_parcels_points.parquet")
fp_remm_parcels   <- here::here(sub_dirs$parcels, "REMM_BY_parcels.parquet")
fp_remm_buildings <- here::here(sub_dirs$parcels, "REMM_BY_buildings.parquet")
```

```{r}
#| label: load-targets
#| description: "Load lightweight targets (Index/Centroid)."

# A. H3: Read Index -> Calc Centroid on-the-fly
#    We ignore the stored polygon geometry to save RAM.
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE points_h3 AS
  SELECT
    row_number() OVER () as oid,
    CAST(h3_index AS UBIGINT) as h3_index,
    ST_Transform(
      ST_Point(
        h3_cell_to_lng(CAST(h3_index AS UBIGINT)),
        h3_cell_to_lat(CAST(h3_index AS UBIGINT))
      ),
      '{CRS_WGS}', '{CRS_UTM}', always_xy := true
    ) as geom
  FROM read_parquet('{fp_h3_grid}')
"))
DBI::dbExecute(conn, "CREATE INDEX idx_h3_utm ON points_h3 USING RTREE (geom)")

# B. Parcels: Read Points directly
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TABLE points_parcels AS
  SELECT
    row_number() OVER () as oid, -- <-- BATCHING
    parcel_id,
    ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
  FROM read_parquet('{fp_parcels_pts}')
"))
DBI::dbExecute(conn, "CREATE INDEX idx_pp_utm ON points_parcels USING RTREE (geom)")
```

```{r}
#| label: detect-res

# AUTO-DETECT RESOLUTION
# We take the first index from the table and ask DuckDB for its resolution.
res <- DBI::dbGetQuery(conn, "SELECT h3_get_resolution(h3_index) as r FROM points_h3 LIMIT 1")$r

if (is.na(res) || length(res) == 0) {
  stop("❌ Could not detect H3 resolution. Is the H3 grid empty?")
}
message(glue::glue("✅ Auto-detected H3 Resolution: {res}"))
```

# Part 2: Static Attributes (Admin & Land Use)

## Building Codes (BC)

We join the REMM buildings table. Since `building_type` contains full strings (e.g., "Single Family Res"), we create a mapping table to translate them to abbreviations (e.g., "SF").

```{r}
#| label: bc-mapping

# Create Mapping Table (Full Name -> Abbreviation)
bc_mapping <- tibble::tribble(
  ~full_name,                  ~abbr,
  "Agriculture",              "AG",
  "Church",                   "CH",
  "Commons",                  "OS",
  "Educational",              "ED",
  "Empty - Buildable",        "EM",
  "Empty Buildable",          "EM",
  "Governmental",             "GO",
  "Group Quarters",           "GQ",
  "Healthcare",               "HE",
  "Industrial",               "IN",
  "Mixed Use",                "MU",
  "Multi Family Res",         "MF",
  "No Build",                 "NB",
  "Office",                   "OF",
  "Open Space Non-Buildable", "OS",
  "Other",                    "OT",
  "Retail",                   "RE",
  "Single Family Res",        "SF",
  "Utilities",                "UT"
  # Note: NA values are handled in the join below
)
DBI::dbWriteTable(conn, "bc_map", bc_mapping, overwrite = TRUE)
```

### BC for Parcels

```{r}
#| label: bc-assign-parcel

# --- A. PARCELS: Attribute Match (ID Join) ---
DBI::dbExecute(conn, glue::glue("
  ALTER TABLE points_parcels ADD COLUMN BC VARCHAR;
  UPDATE points_parcels
  SET BC = COALESCE(mapped.abbr, 'OT')
  FROM (
    SELECT b.parcel_id, m.abbr
    FROM read_parquet('{fp_remm_buildings}') b
    JOIN bc_map m ON b.building_type = m.full_name
    QUALIFY ROW_NUMBER() OVER (PARTITION BY b.parcel_id ORDER BY b.building_sqft DESC) = 1
  ) mapped
  WHERE points_parcels.parcel_id = mapped.parcel_id
"))

DBI::dbExecute(conn, "UPDATE points_parcels SET BC = 'OT' WHERE BC IS NULL")
```

### BC for H3

```{r}
#| label: bc-assign-h3

message("🏗️ Calculating Building Codes (BC) for h3...")

# --- B. H3: Index Match (Direct Building Join) ---
# We calculate the H3 index of every building on the fly to match them to our grid.
DBI::dbExecute(conn, glue::glue("
  ALTER TABLE points_h3 ADD COLUMN BC VARCHAR;

  -- 1. Create Lookup of H3 Indices from Buildings
  CREATE OR REPLACE TEMP TABLE h3_build_map AS
  SELECT
    CAST(h3_latlng_to_cell(ST_Y(ST_Centroid(geometry)), ST_X(ST_Centroid(geometry)), {res}) AS UBIGINT) as h3_index,
    MODE(COALESCE(m.abbr, 'OT')) as mode_bc
  FROM read_parquet('{fp_remm_buildings}') b
  LEFT JOIN bc_map m ON b.building_type = m.full_name
  GROUP BY 1;

  -- 2. Update Target
  UPDATE points_h3
  SET BC = map.mode_bc
  FROM h3_build_map map
  WHERE points_h3.h3_index = map.h3_index;
"))

# --- C. H3: Spatial Gap Fill (Temporary Polygon Load) ---
# We load the heavy polygons ONLY for this specific query, then drop them.
message("  ↳ Filling H3 gaps using Parcel Polygons (Temporary Load)...")

# Join BC from points_parcels to the raw polygon file to create a temp fill layer
DBI::dbExecute(conn, glue::glue("
  CREATE OR REPLACE TEMP TABLE temp_poly_fill AS
  SELECT
    pts.BC,
    ST_Transform(raw.geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
  FROM read_parquet('{fp_remm_parcels}') raw
  JOIN points_parcels pts ON raw.parcel_id = pts.parcel_id
  WHERE pts.BC IS NOT NULL
"))
DBI::dbExecute(conn, "CREATE INDEX idx_temp_poly ON temp_poly_fill USING RTREE (geom)")

# Run the Spatial Backfill
DBI::dbExecute(conn, "
  UPDATE points_h3
  SET BC = matched.BC
  FROM (
    SELECT h.h3_index, p.BC
    FROM points_h3 h
    JOIN temp_poly_fill p ON ST_Intersects(h.geom, p.geom)
    WHERE h.BC IS NULL
  ) matched
  WHERE points_h3.h3_index = matched.h3_index
")

# CLEAN UP IMMEDIATELY
DBI::dbExecute(conn, "DROP TABLE temp_poly_fill")
message("  ↳ Temporary polygons dropped. Memory freed.")

# --- D. Final Cleanup ---
DBI::dbExecute(conn, "UPDATE points_h3 SET BC = 'OT' WHERE BC IS NULL")
DBI::dbExecute(conn, "DROP TABLE bc_map; DROP TABLE h3_build_map;")

message("✅ Targets Ready: Points Loaded, BC Calculated.")
```

```{r}
#| label: check-missing-b3
#| eval: false

# Check H3 for missing BC values
DBI::dbGetQuery(conn, "
  SELECT
    COUNT(*) as total,
    COUNT(BC) as filled,
    COUNT(*) - COUNT(BC) as missing,
    ROUND(100.0 * (COUNT(*) - COUNT(BC)) / COUNT(*), 2) as missing_pct
  FROM points_h3
")

# Check Parcels (for comparison)
DBI::dbGetQuery(conn, "
  SELECT
    COUNT(*) as total,
    COUNT(BC) as filled,
    COUNT(*) - COUNT(BC) as missing,
    ROUND(100.0 * (COUNT(*) - COUNT(BC)) / COUNT(*), 2) as missing_pct
  FROM points_parcels
")
```

## Attach Administrative Data

A helper function to join City/County codes to any target table (Parcel or H3).

```{r}
#| label: func-admin-join

# Updated to work on a SINGLE table (the point table)
attach_admin_data <- function(conn, table_name) {

  message(glue::glue("🏛️ Attaching Admin Data to {table_name}..."))
  fp_municipal_bound <- here::here(sub_dirs$base, "UtahMunicipalBoundaries.parquet")
  fp_county_bound    <- here::here(sub_dirs$base, "UtahCountyBoundaries.parquet")

  # 1. Load Admin Layers (Fixed Geometry)
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_muni AS
    SELECT ugrcode as CommCode, ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_municipal_bound}')
  "))

  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE layer_county AS
    SELECT countynbr as COUNTYNBR, ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true) as geom
    FROM read_parquet('{fp_county_bound}')
  "))

  # 2. Add Columns & Update
  for (col in c("CommCode", "COUNTYNBR")) {
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} DROP COLUMN IF EXISTS {col}"))
    DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN {col} VARCHAR"))
  }

  DBI::dbExecute(conn, glue::glue("
    UPDATE {table_name} SET CommCode = m.CommCode
    FROM layer_muni m WHERE ST_Intersects({table_name}.geom, m.geom)
  "))

  DBI::dbExecute(conn, glue::glue("
    UPDATE {table_name} SET COUNTYNBR = c.COUNTYNBR
    FROM layer_county c WHERE ST_Intersects({table_name}.geom, c.geom)
  "))

  DBI::dbExecute(conn, "DROP TABLE layer_muni; DROP TABLE layer_county;")
}
```

```{r}
#| label: run-admin-join

# Pass the pre-calculated point tables
attach_admin_data(conn, "points_parcels")
attach_admin_data(conn, "points_h3")
```

## Attach Job Access (AA, AT)

We attach the raw job access numbers from the TAZ layer. The normalization (dividing by city max) will happen in Step 4.

```{r}
#| label: func-job-join

fp_taz_access <- here::here(dirs$processed, "access", "TAZ_AccessToOpportunities_OY2023.parquet")

attach_job_access <- function(conn, table_name) {

  message(glue::glue("💼 Attaching Job Access to {table_name}..."))

  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE taz_access AS
    SELECT
      ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)) as geom,
      job_by_auto, job_by_tran
    FROM read_parquet('{fp_taz_access}')
    WHERE ST_IsValid(geometry)
  "))
  DBI::dbExecute(conn, "CREATE INDEX idx_taz_geom ON taz_access USING RTREE (geom)")

  DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN IF NOT EXISTS \"AA\" DOUBLE"))
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN IF NOT EXISTS \"AT\" DOUBLE"))

  DBI::dbExecute(conn, glue::glue("
    UPDATE {table_name}
    SET \"AA\" = t.job_by_auto, \"AT\" = t.job_by_tran
    FROM taz_access t
    WHERE ST_Intersects({table_name}.geom, t.geom)
  "))

  DBI::dbExecute(conn, "DROP TABLE taz_access;")
}
```

```{r}
#| label: run-job-join

# Run with pre-calculated points
attach_job_access(conn, "points_parcels")
attach_job_access(conn, "points_h3")
```

# Part 3: Amenity Operations Engine

## Configuration

We define all external layers here. This structure handles Binary (In/Out), Isochrone (Time), and Buffer (Distance) logic.

```{r}
#| label: config-amenities

# Helpers
iso_walk  <- function(f) here::here(dirs$isochrones, "foot-walking", f)
iso_bike  <- function(f) here::here(dirs$isochrones, "cycling-regular", f)
iso_drive <- function(f) here::here(dirs$isochrones, "driving-car", f)
land_use  <- function(f) here::here(sub_dirs$landuse, f)
trans     <- function(f) here::here(sub_dirs$transport, f)
amen      <- function(f) here::here(sub_dirs$amenities, f)

# Added 'tolerance' column (last column)
amenity_config <- tibble::tribble(
  ~col_name, ~type,       ~path,                                    ~filter_sql,           ~value_col, ~buffer_opts, ~tolerance,

  # --- BINARY (Simple shapes, low tolerance ok) ---
  "OZ",      "binary",    land_use("Utah_Qualified_Opportunity_Zones.parquet"), "1=1",      NA,         NA,       10.0,
  "AP",      "binary",    amen("WalkAccesstoParks_10Min.parquet"),              "1=1",      NA,         NA,       10.0,
  "CM",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Metropolitan Center'", NA, NA, 10.0,
  "CU",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Urban Center'",        NA, NA, 10.0,
  "CC",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'City Center'",         NA, NA, 10.0,
  "CN",      "binary",    land_use("WCV_Centers_and_Regional_Land_Uses.parquet"), "center_type = 'Neighborhood Center'", NA, NA, 10.0,

  # --- ISOCHRONES (Complex shapes) ---
  "TT_Local", "isochrone", iso_walk("Transit_Local.parquet"),        "1=1", "value", NA, 25.0,
  "TT_Rapid", "isochrone", iso_walk("Transit_Rapid.parquet"),        "1=1", "value", NA, 25.0,
  "TT_CRT",   "isochrone", iso_bike("Transit_CRT.parquet"),          "1=1", "value", NA, 25.0,
  "TF",       "isochrone", iso_drive("Freeways.parquet"),            "1=1", "value", NA, 100.0,
  "AC",       "isochrone", iso_walk("ChildCare.parquet"),            "1=1", "value", NA, 25.0,
  "AH",       "isochrone", iso_drive("Healthcare.parquet"),          "1=1", "value", NA, 100.0,
  "AG",       "isochrone", iso_walk("Grocery.parquet"),              "1=1", "value", NA, 25.0,
  "AM",       "isochrone", iso_walk("Community_Centers.parquet"),    "1=1", "value", NA, 25.0,
  "AE_K12",   "isochrone", iso_walk("Schools_PreK12.parquet"),       "1=1", "value", NA, 25.0,
  "AE_High",  "isochrone", iso_bike("Schools_HigherEd.parquet"),     "1=1", "value", NA, 50.0,

  # --- BUFFER (Lines need low tolerance to stay accurate) ---
  "TA_class1", "buffer", trans("Bikeways.parquet"), "facility1 LIKE '%(1A)%' OR facility1 LIKE '%(1B)%'", NA, c(0.25, 0.5, 0.75, 1.5, 3.0), 25.0,
  "TA_class2", "buffer", trans("Bikeways.parquet"), "facility1 LIKE '%(2A)%' OR facility1 LIKE '%(2B)%' OR facility1 LIKE '%2C%'", NA, c(0.25, 0.5, 0.75, 1.5, 3.0), 25.0,
  "TA_trails", "buffer", trans("Bikeways.parquet"), "facility1 LIKE '%Trail%'", NA, c(0.25, 0.5, 0.75, 1.5, 3.0), 25.0
)
```

## Processing Function

### Helper Function

This function handles the transform, union-by-value optimization, and spatial joins entirely in UTM 12N.

```{r}
#| label: func-spatial-engine

# Initialize logger
log_file <- here::here("_output", "log_spatial_progress.txt")
write(paste("Started at:", Sys.time()), file = log_file)

log_msg <- function(msg) {
  message(msg)
  write(paste(Sys.time(), "-", msg), file = log_file, append = TRUE)
}

# -----------------------------------------------------------------------------
# FUNCTION 1: PREPARE (With Timer & Crash Fix)
# -----------------------------------------------------------------------------
prepare_amenity_data <- function(conn, config) {

  t_start <- Sys.time()
  path <- config$path
  type <- config$type
  sql_filter <- config$filter_sql
  val_col <- config$value_col
  tol <- if(!is.null(config$tolerance)) config$tolerance else 10.0

  if (!fs::file_exists(path)) return(FALSE)

  log_msg(glue::glue("📦 Preparing: {config$col_name} (Tol: {tol}m)"))

  DBI::dbExecute(conn, "DROP TABLE IF EXISTS tmp_amenity_stage")

  # 1. Load & Simplify & REPAIR (ST_Buffer 0)
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE tmp_amenity_stage AS
    SELECT
      -- ST_Buffer(
        ST_Simplify(
          ST_MakeValid(ST_Transform(geometry, '{CRS_WGS}', '{CRS_UTM}', always_xy := true)),
          {tol}
      --  ), 0
      ) as geom
      {if(type=='isochrone') glue::glue(', {val_col} as val') else ''}
    FROM read_parquet('{path}')
    WHERE {sql_filter}
  "))

  # 2. Aggregation Strategy
  DBI::dbExecute(conn, "DROP TABLE IF EXISTS tmp_amenity_ready")
  if (type == "isochrone") {
    DBI::dbExecute(conn, "CREATE TEMP TABLE tmp_amenity_ready AS SELECT val, ST_Union_Agg(geom) as geom FROM tmp_amenity_stage GROUP BY val")
  } else {
    DBI::dbExecute(conn, "CREATE TEMP TABLE tmp_amenity_ready AS SELECT * FROM tmp_amenity_stage")
  }

  DBI::dbExecute(conn, "CREATE INDEX idx_amenity_ready ON tmp_amenity_ready USING RTREE (geom)")
  DBI::dbExecute(conn, "DROP TABLE tmp_amenity_stage")

  duration <- round(difftime(Sys.time(), t_start, units = "secs"), 2)
  log_msg(glue::glue("   ✅ Prepared in {duration}s"))
  return(TRUE)
}

# -----------------------------------------------------------------------------
# FUNCTION 2: APPLY (With Timer - No Batching)
# -----------------------------------------------------------------------------
apply_amenity_logic <- function(conn, target_table, config) {

  t_start <- Sys.time()
  col  <- config$col_name
  type <- config$type

  # Reset Column
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} DROP COLUMN IF EXISTS {col}"))
  DBI::dbExecute(conn, glue::glue("ALTER TABLE {target_table} ADD COLUMN {col} DOUBLE DEFAULT NULL"))

  if (type == "binary") {
    DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = 1 WHERE EXISTS (SELECT 1 FROM tmp_amenity_ready r WHERE ST_Intersects({target_table}.geom, r.geom))"))
    DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = 0 WHERE {col} IS NULL"))

  } else if (type == "isochrone") {
    vals <- DBI::dbGetQuery(conn, "SELECT DISTINCT val FROM tmp_amenity_ready ORDER BY val ASC")$val
    for (v in vals) {
      DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = {v} FROM tmp_amenity_ready g WHERE {target_table}.{col} IS NULL AND g.val = {v} AND ST_Intersects({target_table}.geom, g.geom)"))
    }

  } else if (type == "buffer") {
    dists <- sort(unlist(config$buffer_opts), decreasing = FALSE)
    for (miles in dists) {
      meters <- miles * 1609.34
      DBI::dbExecute(conn, glue::glue("UPDATE {target_table} SET {col} = {miles} WHERE {col} IS NULL AND EXISTS (SELECT 1 FROM tmp_amenity_ready r WHERE ST_DWithin({target_table}.geom, r.geom, {meters}))"))
    }
  }

  duration <- round(difftime(Sys.time(), t_start, units = "secs"), 2)
  log_msg(glue::glue("   🚀 Applied to {target_table} in {duration}s"))
}
```

### Spatial Join

We iterate through the config and apply logic to both tables.

```{r}
#| label: run-engine

message("🚀 Starting Staged Processing on Points...")

# 1. Start Total Timer
t_total_start <- Sys.time()

amenity_config |>
  purrr::pwalk(function(...) {

    # Capture the row data as a list
    row <- list(...)

    # 1. Prepare (Once)
    success <- tryCatch({
      prepare_amenity_data(conn, row)
    }, error = function(e) {
      log_msg(glue::glue("❌ ERROR Preparing {row$col_name}: {e$message}"))
      return(FALSE)
    })

    # 2. Apply (To Targets)
    if (success) {
      tryCatch({
        apply_amenity_logic(conn, "points_parcels", row)
        apply_amenity_logic(conn, "points_h3", row)
      }, error = function(e) {
        log_msg(glue::glue("❌ ERROR Applying {row$col_name}: {e$message}"))
      })

      DBI::dbExecute(conn, "DROP TABLE IF EXISTS tmp_amenity_ready")
      gc()
    }
  })

# 4. End Total Timer & Log
t_total_end <- Sys.time()
total_duration <- round(difftime(t_total_end, t_total_start, units = "mins"), 2)

message(glue::glue("✅ Points enriched successfully. Total Time: {total_duration} mins."))
```

# Export

We export the "Wide" tables (ID + all amenity times). We exclude the geometry to keep the files lightweight for the next scoring step.

```{r}
#| label: export

# 1. Parcels
fp_out_parcels <- here::here(dirs$results, "parcels_pt_enriched.parquet")
DBI::dbExecute(conn, glue::glue("
  COPY (
    SELECT * EXCLUDE(geom), ST_Transform(geom, '{CRS_UTM}', '{CRS_WGS}', always_xy := true) AS geom
    FROM points_parcels
  ) TO '{fp_out_parcels}' (FORMAT PARQUET)
"))

# 2. H3
fp_out_h3 <- here::here(dirs$results, "h3_pt_enriched.parquet")
DBI::dbExecute(conn, glue::glue("
  COPY (
    SELECT * EXCLUDE(geom), ST_Transform(geom, '{CRS_UTM}', '{CRS_WGS}', always_xy := true) AS geom
    FROM points_h3
  ) TO '{fp_out_h3}' (FORMAT PARQUET)
"))

message("🚀 Processing Complete!")
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
