---
title: ATO Scoring & Normalization
subtitle: Calculate Final Housing Suitability Component Scores
description: Normalize raw metric values into 0-1 scores based on city maximums and proximity decay functions.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-15"
---

# Environment Setup

## Libraries

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(DBI)
library(duckdb)
library(arrow)
library(geoarrow)
library(glue)
library(here)
library(fs)
```

# Configuration

```{r}
#| label: setup-dirs

dirs <- list(
  results = here::here("_output")
)

# Input Paths (From Step 3)
fp_parcels_enriched <- here::here(dirs$results, "parcels_enriched.parquet")
fp_h3_enriched      <- here::here(dirs$results, "h3_enriched.parquet")

# Output Paths
fp_parcels_scored   <- here::here(dirs$results, "parcels_scored.parquet")
fp_h3_scored        <- here::here(dirs$results, "h3_scored.parquet")
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Initialize Database

```{r}
#| label: init-db
conn <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
```

# Scoring & Weighting

## Scoring Logic

We define helper functions to standardize the scoring logic across both Parcels and H3 grids.

1. Identifies "static" columns (IDs, Admin codes) vs "scorable" columns.
2. Calculates the global MAX or MIN for normalization.
3. Applies **Inverse Normalization** ($\frac{Min}{Value}$) for Cost metrics (Time/Distance).
4. Applies **Linear Normalization** ($\frac{Value}{Max}$) for Benefit metrics (Jobs/Binary).
5. Exports the final table, replacing raw values with scores.

This function encapsulates the entire scoring workflow so it can be applied identically to both Parcels and H3 grids.

```{r}
#| label: func-scoring

score_and_export <- function(conn, input_path, output_path, table_name) {

  message(glue::glue("📊 Loading, Scoring and Normalizing {table_name}..."))

  # 2. Load Data
  DBI::dbExecute(conn, as.character(glue::glue(
    "CREATE OR REPLACE TEMP TABLE {table_name} AS
     SELECT * FROM read_parquet('{input_path}')"
  )))

  # 3. Define Static Columns (SKIP these)
  static_cols <- c("rowid", "geometry", "geom", "parcel_id", "h3_index",
                   "BC", "CommCode", "city", "county", "fips", "COUNTYNBR")

  # 4. Identify Scorable Columns
  all_cols <- DBI::dbListFields(conn, table_name)
  score_cols <- setdiff(all_cols, static_cols)

  if(length(score_cols) == 0) {
    stop("❌ Error: No scorable columns found!")
  }

  # 5. Iterate and Score
  for (col in score_cols) {

    # 🛑 CRITICAL FIX: Quote "{col}" to handle reserved keywords like "AT"
    DBI::dbExecute(conn, glue::glue('ALTER TABLE {table_name} ALTER COLUMN "{col}" TYPE DOUBLE'))

    # ---------------------------------------------------------
    # LOGIC SWITCH
    # ---------------------------------------------------------

    # COST METRICS (Lower is Better -> Min / Value)
    if (col %in% c("TA", "TF") || grepl("^(TT_|AC|AH|AG|AM|AE_)", col)) {

      # Quote "{col}" in SELECT
      sql_min <- as.character(glue::glue('SELECT MIN("{col}") as m FROM {table_name} WHERE "{col}" > 0'))
      min_val <- DBI::dbGetQuery(conn, sql_min)$m
      if (is.na(min_val)) min_val <- 1.0

      message(glue::glue("   ⚖️ Scoring {col} (Inverse: Min/Val, Best={round(min_val, 2)})..."))

      # Quote "{col}" in UPDATE
      DBI::dbExecute(conn, as.character(glue::glue('
        UPDATE {table_name}
        SET "{col}" = CASE
          WHEN "{col}" IS NULL THEN 0
          WHEN "{col}" <= {min_val} THEN 1.0
          ELSE {min_val} / "{col}"
        END
      ')))

    # BENEFIT METRICS (Higher is Better -> Value / Max)
    } else {

      sql_max <- as.character(glue::glue('SELECT MAX("{col}") as m FROM {table_name}'))
      max_val <- DBI::dbGetQuery(conn, sql_max)$m
      if (is.na(max_val) || max_val == 0) max_val <- 1.0

      message(glue::glue("   ⚖️ Scoring {col} (Linear: Val/Max, Best={round(max_val, 2)})..."))

      DBI::dbExecute(conn, as.character(glue::glue('
        UPDATE {table_name}
        SET "{col}" = CASE
          WHEN "{col}" IS NULL THEN 0
          ELSE "{col}" / {max_val}
        END
      ')))
    }
  }

  # -------------------------------------------------------------------------
  # 6. Composite Scores
  # -------------------------------------------------------------------------
  message("   🔗 Calculating Composite Scores...")

  # Transit Score (TT)
  if (all(c("TT_Local", "TT_Rapid", "TT_CRT") %in% score_cols)) {
    DBI::dbExecute(conn, glue::glue("
      ALTER TABLE {table_name} DROP COLUMN IF EXISTS TT;
      ALTER TABLE {table_name} ADD COLUMN TT DOUBLE;
      UPDATE {table_name}
      SET TT = GREATEST(
        COALESCE(TT_Local, 0) * 0.5,
        COALESCE(TT_Rapid, 0) * 0.8,
        COALESCE(TT_CRT,   0) * 1.0
      )
    "))
  }

  # Education Score (AE)
  if (all(c("AE_K12", "AE_High") %in% score_cols)) {
    DBI::dbExecute(conn, glue::glue("
      ALTER TABLE {table_name} DROP COLUMN IF EXISTS AE;
      ALTER TABLE {table_name} ADD COLUMN AE DOUBLE;
      UPDATE {table_name}
      SET AE = GREATEST(
        COALESCE(AE_K12,  0) * 0.8,
        COALESCE(AE_High, 0) * 1.0
      )
    "))
  }

  # 7. Export
  fs::dir_create(fs::path_dir(output_path))
  message(glue::glue("💾 Saving Scored Data to {output_path}..."))

  DBI::dbExecute(conn, as.character(glue::glue("
    COPY (
      SELECT * FROM {table_name}
    ) TO '{output_path}' (FORMAT PARQUET)
  ")))

  DBI::dbExecute(conn, as.character(glue::glue("DROP TABLE {table_name}")))
}
```

## Run Scoring

### Parcels

```{r}
#| label: run-parcels
score_and_export(conn, fp_parcels_enriched, fp_parcels_scored, "targets_parcels")
```

## H3 Grid

```{r}
#| label: run-h3
score_and_export(conn, fp_h3_enriched, fp_h3_scored, "targets_h3")
```

### Export to GDB

```{r}
#| label: export-gdb

message("🌍 Converting Parquets to File Geodatabase...")

# 1. Configuration
gdb_path <- here::here(dirs$results, "HousingSuitability_Centers2025.gdb")

# Define inputs: Layer Name -> File Path
to_export <- list(
  "Parcels_Enriched" = fp_parcels_enriched,
  "H3_Enriched"      = fp_h3_enriched,
  "Parcels_Scored"   = fp_parcels_scored,
  "H3_Scored"        = fp_h3_scored
)

# 2. Execution
purrr::iwalk(to_export, function(path, layer_name) {

  if (!fs::file_exists(path)) return(message(glue::glue("⚠️ Skipping {layer_name} (Not found)")))

  message(glue::glue("  ↳ Processing {layer_name}..."))

  # Streamlined Approach: Open -> Convert to SF -> Write
  arrow::open_dataset(path) |>
    sf::st_as_sf(crs = 4326) |> # WGS84 (matches your source)
    sf::st_cast("MULTIPOLYGON") |>
    # Optional: Uncomment next line if you need the GDB in UTM (Standard for WFRC)
    # sf::st_transform(26912) |>
    sf::st_write(dsn = gdb_path, layer = layer_name, driver = "OpenFileGDB",
                 append = FALSE, quiet = TRUE)
})

message("✅ GDB Export Complete!")
```

## Validation

Check a few records to ensure scores make sense (e.g., scores shouldn't be > 1.0 or infinite).

```{r}
#| label: validate
#| echo: false

arrow::open_dataset(fp_parcels_scored) |> head() |> collect()
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
