---
title: ATO Scoring & Normalization
subtitle: Calculate Final Housing Suitability Component Scores
description: Normalize raw metric values into 0-1 scores based on city maximums and proximity decay functions.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Libraries

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(DBI)
library(duckdb)
library(glue)
library(here)
library(fs)
```

# Configuration

```{r}
#| label: setup-dirs

dirs <- list(
  results = here::here("_output")
)

# Input Paths (From Step 3)
fp_parcels_enriched <- here::here(dirs$results, "parcels_enriched.parquet")
fp_h3_enriched      <- here::here(dirs$results, "h3_enriched.parquet")

# Output Paths
fp_parcels_scored   <- here::here(dirs$results, "parcels_scored.parquet")
fp_h3_scored        <- here::here(dirs$results, "h3_scored.parquet")
```

## Initialize Database

```{r}
#| label: init-db
conn <- DBI::dbConnect(duckdb::duckdb())
```

# Scoring Logic

We define helper functions to standardize the scoring logic across both Parcels and H3 grids.

* **Logic:** `Score = Benchmark / Actual`.
* **Benchmark:** Determined dynamically (Minimum observed value).
* **Cap:** If `Actual <= Benchmark`, Score is **1.0**.

```{r}
#| label: define-macros

# Define a scoring function for inverse decay
# Usage: score_decay(col_name, best_val)
# If NULL (no access), score is 0.
query_macro <- "
CREATE OR REPLACE MACRO score_decay(val, best) AS
CASE
    WHEN val IS NULL THEN 0
    WHEN val < best THEN 1.0  -- Cap score at 1.0 if closer or equal to benchmark
    ELSE ROUND(best / val, 4)
END;
"
DBI::dbExecute(conn, query_macro)
```

# Scoring & Weighting

This function encapsulates the entire scoring workflow so it can be applied identically to both Parcels and H3 grids.

```{r}
#| label: func-scoring

process_scoring <- function(conn, input_path, output_path, table_name) {

  message(glue::glue("🚀 Scoring {table_name}..."))

  # 1. Load Data
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TABLE {table_name} AS SELECT * FROM read_parquet('{input_path}')
  "))

  # ===========================================================================
  # A. JOB ACCESS SCORING (Normalized by City Max)
  #    Logic: Score = Value / Max_Value_In_City
  # ===========================================================================
  message("   📊 Normalizing Job Access by City (CommCode)...")

  cols_in_table <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field

  # Only run if CommCode exists (Parcels/H3 should both have it now)
  if("CommCode" %in% cols_in_table) {
    DBI::dbExecute(conn, glue::glue("
      CREATE OR REPLACE TEMP TABLE job_scores AS
      SELECT
        rowid,
        CASE
          WHEN MAX(AA) OVER (PARTITION BY CommCode) = 0 THEN 0
          ELSE ROUND(AA / MAX(AA) OVER (PARTITION BY CommCode), 4)
        END as AA_score,
        CASE
          WHEN MAX(AT) OVER (PARTITION BY CommCode) = 0 THEN 0
          ELSE ROUND(AT / MAX(AT) OVER (PARTITION BY CommCode), 4)
        END as AT_score
      FROM {table_name}
    "))

    DBI::dbExecute(conn, glue::glue("
      ALTER TABLE {table_name} ADD COLUMN AA_score DOUBLE;
      ALTER TABLE {table_name} ADD COLUMN AT_score DOUBLE;

      UPDATE {table_name}
      SET
        AA_score = j.AA_score,
        AT_score = j.AT_score
      FROM job_scores j
      WHERE {table_name}.rowid = j.rowid
    "))
    DBI::dbExecute(conn, "DROP TABLE job_scores")
  }

  # ===========================================================================
  # B. DYNAMIC BENCHMARKING (Minimum Value)
  #    We calculate the 'Best' (Minimum) value for every amenity column.
  # ===========================================================================
  message("   ⏱️  Calculating Dynamic Benchmarks (Minimum)...")

  target_cols <- c(
    # Transit Components (Time)
    "TT_Local", "TT_Rapid", "TT_CRT",
    # Education Components (Time)
    "AE_K12", "AE_High",
    # Other Isochrones (Time)
    "TF", "AG", "AC", "AH", "AM",
    # Buffers (Distance)
    "TA"
  )

  # Filter to columns that actually exist in the table
  available_cols <- intersect(target_cols, cols_in_table)

  # Calculate MINIMUM (non-NA) for each available column [cite: 104]
  if(length(available_cols) > 0) {

    stats_sql <- paste0(
      "SELECT ",
      paste0(
        glue::glue("MIN({available_cols}) AS {available_cols}_best"),
        collapse = ", "
      ),
      glue::glue(" FROM {table_name}")
    )

    benchmarks <- DBI::dbGetQuery(conn, stats_sql)

    print("   ✅ Dynamic Benchmarks (Best Observed Values):")
    print(benchmarks)

    # ===========================================================================
    # C. CALCULATE RAW DECAY SCORES
    # ===========================================================================

    for (col in available_cols) {
      best_val <- benchmarks[[paste0(col, "_best")]]

      # Safety check: If a column is all NULL, MIN returns NA. Skip scoring.
      if (!is.na(best_val)) {

        # Create "_raw" columns to hold the intermediate score
        new_col <- paste0(col, "_raw")

        DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN IF NOT EXISTS {new_col} DOUBLE"))

        # Apply Decay Macro: Score = best / val
        DBI::dbExecute(conn, glue::glue("UPDATE {table_name} SET {new_col} = score_decay({col}, {best_val})"))
      }
    }
  }

  # ===========================================================================
  # D. COMPOSITE SCORING
  # ===========================================================================

  # 1. TRANSIT (TT)
  #    Max( Local*0.5, Rapid*0.8, CRT*1.0 ) [cite: 108]
  message("   🚌 Calculating Composite Transit Score (TT)...")

  # Helper to safe-reference columns (COALESCE to 0 if they don't exist)
  safe_col <- function(cname) {
    if (paste0(cname, "_raw") %in% DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field) {
      return(paste0("COALESCE(", cname, "_raw, 0)"))
    } else {
      return("0")
    }
  }

  transit_sql <- glue::glue("
    ALTER TABLE {table_name} ADD COLUMN TT_score DOUBLE;
    UPDATE {table_name}
    SET TT_score = GREATEST(
      {safe_col('TT_Local')} * 0.5,
      {safe_col('TT_Rapid')} * 0.8,
      {safe_col('TT_CRT')}   * 1.0
    )
  ")
  DBI::dbExecute(conn, transit_sql)

  # 2. EDUCATION (AE)
  #    Max( K12, HigherEd ) [cite: 109]
  message("   🎓 Calculating Composite Education Score (AE)...")

  edu_sql <- glue::glue("
    ALTER TABLE {table_name} ADD COLUMN AE_score DOUBLE;
    UPDATE {table_name}
    SET AE_score = GREATEST(
      {safe_col('AE_K12')},
      {safe_col('AE_High')}
    )
  ")
  DBI::dbExecute(conn, edu_sql)

  # 3. STANDARD LAYERS (Pass through _raw to _score)
  #    Grocery (AG), Health (AH), Childcare (AC), CommCtr (AM), Freeway (TF), ActiveTrans (TA)

  standard_map <- c(
    "AG_score" = "AG_raw",
    "AH_score" = "AH_raw",
    "AC_score" = "AC_raw",
    "AM_score" = "AM_raw",
    "TF_score" = "TF_raw",
    "TA_score" = "TA_raw"
  )

  for (final in names(standard_map)) {
    raw <- standard_map[[final]]
    # Only update if the raw column was actually created
    if(paste0(raw) %in% DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field) {
      DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN IF NOT EXISTS {final} DOUBLE"))
      DBI::dbExecute(conn, glue::glue("UPDATE {table_name} SET {final} = {raw}"))
    }
  }

  # ===========================================================================
  # E. BINARY COLUMNS (Cleanup)
  # ===========================================================================
  binary_cols <- c("OZ", "AP", "CM", "CU", "CC", "CN")
  for (col in binary_cols) {
     if (col %in% cols_in_table) {
       DBI::dbExecute(conn, glue::glue("UPDATE {table_name} SET {col} = 0 WHERE {col} IS NULL"))
     }
  }

  # ===========================================================================
  # F. EXPORT
  # ===========================================================================
  message(glue::glue("💾 Saving results to {output_path}"))

  # Exclude raw intermediate columns
  all_cols <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field
  raw_cols <- all_cols[grepl("_raw$", all_cols)]

  if(length(raw_cols) > 0) {
    exclude_clause <- paste(raw_cols, collapse = ", ")
    query <- glue::glue("COPY (SELECT * EXCLUDE ({exclude_clause}) FROM {table_name}) TO '{output_path}' (FORMAT PARQUET)")
  } else {
    query <- glue::glue("COPY {table_name} TO '{output_path}' (FORMAT PARQUET)")
  }

  DBI::dbExecute(conn, query)
}
```

# Run Scoring

## Parcels

```{r}
#| label: run-parcels
process_scoring(conn, fp_parcels_enriched, fp_parcels_scored, "parcels_scored")
```

## H3 Grid

```{r}
#| label: run-h3
process_scoring(conn, fp_h3_enriched, fp_h3_scored, "h3_scored")
```

## Validation

Check a few records to ensure scores make sense (e.g., scores shouldn't be > 1.0 or infinite).

```{r}
#| label: validate
#| echo: false

cat("--- Score Summary ---\n")
DBI::dbGetQuery(conn, "
  SELECT
    'Parcels' as layer,
    AVG(TT_score) as mean_Transit,
    AVG(AE_score) as mean_Education,
    AVG(AG_score) as mean_Grocery,
    AVG(TA_score) as mean_Bikeways
  FROM parcels_scored
  UNION ALL
  SELECT
    'H3' as layer,
    AVG(TT_score) as mean_Transit,
    AVG(AE_score) as mean_Education,
    AVG(AG_score) as mean_Grocery,
    AVG(TA_score) as mean_Bikeways
  FROM h3_scored
") |> print()
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
