---
title: ATO Scoring & Normalization
subtitle: Calculate Final Housing Suitability Component Scores
description: Normalize raw metric values into 0-1 scores based on city maximums and proximity decay functions.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-15"
---

# Environment Setup

## Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here",        # https://github.com/r-lib/here/
    "fs",          # https://github.com/r-lib/fs/
    "zip"          # https://github.com/r-lib/zip
    ),
    dependencies = TRUE
  )
```

OR,

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak("renv", dependencies = TRUE) # https://github.com/rstudio/renv/

renv::restore()
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(DBI)
library(duckdb)
library(arrow)
library(geoarrow)
library(glue)
library(here)
library(fs)
```

# Configuration

```{r}
#| label: setup-dirs

dirs <- list(
  results = here::here("_output")
)

# Input Paths (From Step 3)
fp_parcels_enriched <- here::here(dirs$results, "parcels_enriched.parquet")
fp_h3_enriched <- here::here(dirs$results, "h3_enriched.parquet")

# Output Paths
fp_parcels_scored <- here::here(dirs$results, "parcels_scored.parquet")
fp_h3_scored <- here::here(dirs$results, "h3_scored.parquet")
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566" # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326" # WGS84 (Geographic - Lat/Lon)
```

## Initialize Database

```{r}
#| label: init-db
conn <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "INSTALL h3 FROM community; LOAD h3;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")
```

# Scoring & Weighting

## Scoring Logic

We define helper functions to standardize the scoring logic across both Parcels and H3 grids.

1. Identifies "static" columns (IDs, Admin codes) vs "scorable" columns.
2. Calculates the global MAX or MIN for normalization.
3. Applies **Inverse Normalization** ($\frac{Min}{Value}$) for Cost metrics (Time/Distance).
4. Applies **Linear Normalization** ($\frac{Value}{Max}$) for Benefit metrics (Jobs/Binary).
5. Exports the final table, replacing raw values with scores.

This function encapsulates the entire scoring workflow so it can be applied identically to both Parcels and H3 grids.

```{r}
#| label: func-scoring

score_and_export <- function(conn, input_path, output_path, table_name) {
  message(glue::glue("📊 Scoring {table_name}..."))

  # 1. Load Data into View (Zero Copy)
  DBI::dbExecute(
    conn,
    glue::glue(
      "
    CREATE OR REPLACE VIEW raw_{table_name} AS SELECT * FROM read_parquet('{input_path}')
  "
    )
  )

  # 2. Identify Columns
  static_cols <- c(
    "rowid",
    "geometry",
    "geom",
    "parcel_id",
    "h3_index",
    "CommCode",
    "COUNTYNBR",
    "BC"
  )

  all_cols <- DBI::dbListFields(conn, glue::glue("raw_{table_name}"))
  score_cols <- setdiff(all_cols, static_cols)

  if (length(score_cols) == 0) {
    stop("❌ Error: No scorable columns found!")
  }

  # 3. Pass 1: Gather Normalization Stats
  #    We build the SQL to normalize individual columns (0 to 1)
  sql_parts <- list()

  message("   ⚖️ Calculating Min/Max statistics...")

  for (col in score_cols) {
    # Detect Type: Cost (Min/Val) vs Benefit (Val/Max)
    # Added 'TA_' explicitly to cost metrics based on your buffer config
    is_cost <- grepl("^(TT_|AC|AH|AG|AM|AE_|TA_|TF)", col)

    if (is_cost) {
      # COST: Inverse Normalization
      val <- DBI::dbGetQuery(
        conn,
        glue::glue(
          'SELECT MIN("{col}") as m FROM raw_{table_name} WHERE "{col}" > 0'
        )
      )$m
      if (is.na(val) || val == 0) {
        val <- 1.0
      } # Safety

      sql_parts[[col]] <- glue::glue(
        '
        CASE
          WHEN "{col}" IS NULL THEN 0.0
          WHEN "{col}" <= {val} THEN 1.0
          ELSE {val} / "{col}"
        END AS "{col}"
      '
      )
    } else {
      # BENEFIT: Linear Normalization
      val <- DBI::dbGetQuery(
        conn,
        glue::glue('SELECT MAX("{col}") as m FROM raw_{table_name}')
      )$m
      if (is.na(val) || val == 0) {
        val <- 1.0
      }

      sql_parts[[col]] <- glue::glue(
        '
        CASE
          WHEN "{col}" IS NULL THEN 0.0
          ELSE "{col}" / {val}
        END AS "{col}"
      '
      )
    }
  }

  # 4. Define Composite Calculations
  #    These logic blocks run on the ALREADY normalized values from step 3.
  #    Using COALESCE(x, 0) ensures we don't get NULLs if one component is missing.

  composites <- list()

  # A. Transit: Weighted Max of normalized Local, Rapid, and CRT
  if (all(c("TT_Local", "TT_Rapid", "TT_CRT") %in% score_cols)) {
    composites[[
      "TT"
    ]] <- "GREATEST(COALESCE(TT_Local,0)*0.5, COALESCE(TT_Rapid,0)*0.8, COALESCE(TT_CRT,0)*1.0) AS TT"
  }

  # B. Education: Weighted Max of K12 and HigherEd
  if (all(c("AE_K12", "AE_High") %in% score_cols)) {
    composites[[
      "AE"
    ]] <- "GREATEST(COALESCE(AE_K12,0)*0.8, COALESCE(AE_High,0)*1.0) AS AE"
  }

  # C. Active Transport (Bikeways): Weighted Max of normalized buffer scores
  if (all(c("TA_class1", "TA_class2", "TA_trails") %in% score_cols)) {
    composites[[
      "TA"
    ]] <- "GREATEST(COALESCE(TA_class2,0)*0.5, COALESCE(TA_class1,0)*1.0, COALESCE(TA_trails,0)*1.0) AS TA"
  }

  # 5. Pass 2: Execute Pipeline
  message(glue::glue("💾 Projecting & Saving to {output_path}..."))

  target_static_cols <- intersect(all_cols, static_cols)

  select_static_list <- lapply(target_static_cols, function(col) {
    if (col == "h3_index") {
      # Convert UBIGINT to Hex String
      return("h3_h3_to_string(h3_index) AS h3_index")
    } else {
      return(col)
    }
  })

  select_static <- paste(unlist(select_static_list), collapse = ", ")
  select_scores <- paste(unlist(sql_parts), collapse = ", \n      ")
  select_comps <- if (length(composites) > 0) {
    paste(",", paste(unlist(composites), collapse = ", \n      "))
  } else {
    ""
  }

  # The CTE 'scored_base' normalizes the data.
  # The outer SELECT calculates composites on those normalized values.
  query <- glue::glue(
    "
    COPY (
      WITH scored_base AS (
        SELECT
          {select_static},
          {select_scores}
        FROM raw_{table_name}
      )
      SELECT
        *
        {select_comps}
      FROM scored_base
    ) TO '{output_path}' (FORMAT PARQUET)
  "
  )

  DBI::dbExecute(conn, query)
  DBI::dbExecute(conn, glue::glue("DROP VIEW raw_{table_name}"))
}
```

## Run Scoring

### Parcels

```{r}
#| label: run-parcels
score_and_export(
  conn,
  fp_parcels_enriched,
  fp_parcels_scored,
  "targets_parcels"
)
```

## H3 Grid

```{r}
#| label: run-h3
score_and_export(conn, fp_h3_enriched, fp_h3_scored, "targets_h3")
```

### Export to GDB

```{r}
#| label: export-gdb

message("🌍 Converting Parquets to File Geodatabase...")

# 1. Configuration
gdb_path <- here::here(dirs$results, "HousingSuitability_Centers202512.gdb")

# Define inputs: Layer Name -> File Path
to_export <- list(
  "Parcels_Enriched" = fp_parcels_enriched,
  "H3_Enriched" = fp_h3_enriched,
  "Parcels_Scored" = fp_parcels_scored,
  "H3_Scored" = fp_h3_scored
)

# 2. Execution
purrr::iwalk(to_export, function(path, layer_name) {
  if (!fs::file_exists(path)) {
    return(message(glue::glue("⚠️ Skipping {layer_name} (Not found)")))
  }

  message(glue::glue("  ↳ Processing {layer_name}..."))

  # Streamlined Approach: Open -> Convert to SF -> Write
  arrow::open_dataset(path) |>
    sf::st_as_sf(crs = CRS_WGS) |> # WGS84
    sf::st_cast("MULTIPOLYGON") |>
    # Optional: Uncomment next line if you need the GDB in UTM
    # sf::st_transform(CRS_UTM) |>
    sf::st_write(
      dsn = gdb_path,
      layer = layer_name,
      driver = "OpenFileGDB",
      append = FALSE,
      quiet = TRUE
    )
})

message("✅ GDB Export Complete!")
```

```{r}
#| label: zip-gdb

message("📦 Compressing GDB to Zip...")

# Define Paths
zip_path <- paste0(gdb_path, ".zip")

# 3. Create Zip
zip::zip(
  zipfile = zip_path, # Where to save the result (Absolute path works best)
  files = fs::path_file(gdb_path), # What to grab (Relative to 'root')
  root = fs::path_dir(gdb_path), # The virtual working directory for this operation
  mode = "mirror" # "mirror" is best for exact copies
)

message(glue::glue("✅ Compressed successfully: {zip_path}"))
```

## Validation

Check a few records to ensure scores make sense (e.g., scores shouldn't be > 1.0 or infinite).

```{r}
#| label: validate
#| echo: false

arrow::open_dataset(fp_parcels_scored) |> head() |> collect()
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
