---
title: ATO Scoring & Normalization
subtitle: Calculate Final Housing Suitability Component Scores
description: Normalize raw metric values into 0-1 scores based on city maximums and proximity decay functions.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-02"
---

# Environment Setup

## Libraries

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(DBI)
library(duckdb)
library(glue)
library(here)
library(fs)
```

# Configuration

```{r}
#| label: setup-dirs

dirs <- list(
  results = here::here("_output")
)

# Input Paths (From Step 3)
fp_parcels_enriched <- here::here(dirs$results, "parcels_enriched.parquet")
fp_h3_enriched      <- here::here(dirs$results, "h3_enriched.parquet")

# Output Paths
fp_parcels_scored   <- here::here(dirs$results, "parcels_scored.parquet")
fp_h3_scored        <- here::here(dirs$results, "h3_scored.parquet")
```

## Initialize Database

```{r}
#| label: init-db
conn <- DBI::dbConnect(duckdb::duckdb())
```

# Scoring Logic

We define helper functions to standardize the scoring logic across both Parcels and H3 grids.

1. Job Access Normalization (City-Based)
*Logic: `Score = Value / Max_Value_In_City`*
* If a city has low job access overall, the best parcel in that city still gets a 1.0.
* We use DuckDB window functions: MAX(val) OVER (PARTITION BY CommCode).

2. Proximity Decay (Isochrones & Buffers)
*Logic: `Score = Best_Threshold / Actual_Value`*

* Example (Walking): Best threshold is 5 mins (300s).
    * If parcel is 300s away: `300 / 300 = 1.0`
    * If parcel is 600s away: `300 / 600 = 0.5`
    * If parcel is 900s away: `300 / 900 = 0.33`
* This matches the Python logic: `1 / (Distance * (1/Smallest))`

```{r}
#| label: define-macros

# Define a scoring function for inverse decay
# Usage: score_decay(col_name, best_val)
# If NULL (no access), score is 0.
query_macro <- "
CREATE OR REPLACE MACRO score_decay(val, best) AS
CASE
    WHEN val IS NULL THEN 0
    WHEN val < best THEN 1.0  -- Cap score at 1.0 if closer than threshold
    ELSE ROUND(best / val, 4)
END;
"
DBI::dbExecute(conn, query_macro)
```

# Execution Function

This function applies the scoring logic to any target table (Parcels or H3).

```{r}
#| label: func-scoring

process_scoring <- function(conn, input_path, output_path, table_name) {

  message(glue::glue("🚀 Scoring {table_name}..."))

  # 1. Load Data
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TABLE {table_name} AS SELECT * FROM read_parquet('{input_path}')
  "))

  # 2. Calculate Job Access Scores (City Normalized)
  #    We create new columns AA_score and AT_score
  #    Handle division by zero if a city has 0 max jobs

  message("   📊 Normalizing Job Access by City (CommCode)...")
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE job_scores AS
    SELECT
      rowid,
      -- Auto Access (AA)
      CASE
        WHEN MAX(AA) OVER (PARTITION BY CommCode) = 0 THEN 0
        ELSE ROUND(AA / MAX(AA) OVER (PARTITION BY CommCode), 4)
      END as AA_score,

      -- Transit Access (AT)
      CASE
        WHEN MAX(AT) OVER (PARTITION BY CommCode) = 0 THEN 0
        ELSE ROUND(AT / MAX(AT) OVER (PARTITION BY CommCode), 4)
      END as AT_score
    FROM {table_name}
  "))

  # Update main table
  DBI::dbExecute(conn, glue::glue("
    ALTER TABLE {table_name} ADD COLUMN AA_score DOUBLE;
    ALTER TABLE {table_name} ADD COLUMN AT_score DOUBLE;

    UPDATE {table_name}
    SET
      AA_score = j.AA_score,
      AT_score = j.AT_score
    FROM job_scores j
    WHERE {table_name}.rowid = j.rowid
  "))
  DBI::dbExecute(conn, "DROP TABLE job_scores")

  # 3. Calculate Decay Scores (Isochrones & Buffers)
  #    We define the 'Best' threshold for each layer below.

  message("   ⏱️  Calculating Proximity Decay Scores...")

  # Configuration mapping: Column Name -> Best Threshold
  # Time is in Seconds (300 = 5 min), Distance in Miles (0.25)
  # NOTE: Ensure 'TA' (Bikeways) matches the units produced in Step 3.
  # If Step 3 output TA in 'miles', use 0.25. If 'meters', use 402.3.
  # Based on Step 3 script, TA buffer loop put 'miles' value into the column.

  scoring_map <- list(
    # Time-Based (Seconds) - Best is 5 mins (300s)
    "TT" = 300,  # Transit
    "TF" = 300,  # Freeway (using transit layer as proxy per prompt?)
    "AC" = 300,  # Childcare
    "AG" = 300,  # Grocery
    "AM" = 300,  # Community Centers
    "AE" = 300,  # Education (Combined)
    "AH" = 300,  # Healthcare

    # Distance-Based (Miles) - Best is 0.25 miles
    "TA" = 0.25  # Bikeways
  )

  for (col in names(scoring_map)) {
    best_val <- scoring_map[[col]]
    new_col <- paste0(col, "_score")

    # Check if column exists before trying to score it
    cols_exist <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field

    if (col %in% cols_exist) {
      # Add column
      DBI::dbExecute(conn, glue::glue("ALTER TABLE {table_name} ADD COLUMN IF NOT EXISTS {new_col} DOUBLE"))

      # Update with Macro
      DBI::dbExecute(conn, glue::glue("
        UPDATE {table_name}
        SET {new_col} = score_decay({col}, {best_val})
      "))
    }
  }

  # 4. Binary Columns (Pass-through)
  #    Ensure 0/1 columns (OZ, AP, Centers) are clean.
  #    We rename them to *_score for consistency if desired,
  #    or just ensure NULLs are 0.

  binary_cols <- c("OZ", "AP", "CM", "CU", "CC", "CN")

  for (col in binary_cols) {
     cols_exist <- DBI::dbGetQuery(conn, glue::glue("DESCRIBE {table_name}"))$Field
     if (col %in% cols_exist) {
       DBI::dbExecute(conn, glue::glue("UPDATE {table_name} SET {col} = 0 WHERE {col} IS NULL"))
     }
  }

  # 5. Export
  message(glue::glue("💾 Saving results to {output_path}"))
  DBI::dbExecute(conn, glue::glue("
    COPY {table_name} TO '{output_path}' (FORMAT PARQUET)
  "))
}
```

# Run Scoring

## Parcels

```{r}
#| label: run-parcels
process_scoring(conn, fp_parcels_in, fp_parcels_out, "parcels_scored")
```

## H3 Grid

```{r}
#| label: run-h3
process_scoring(conn, fp_h3_in, fp_h3_out, "h3_scored")
```

## Validation

Check a few records to ensure scores make sense (e.g., scores shouldn't be > 1.0 or infinite).

```{r}
#| label: validate
#| echo: false

cat("--- Parcel Score Summary ---\n")
DBI::dbGetQuery(conn, "
  SELECT
    AVG(AA_score) as mean_AA,
    AVG(AG_score) as mean_Grocery,
    AVG(TT_score) as mean_Transit,
    MAX(AG_score) as max_Grocery
  FROM parcels_scored
") |> print()
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
