---
title: ATO Scoring & Normalization
subtitle: Calculate Final Housing Suitability Component Scores
description: Normalize raw metric values into 0-1 scores based on city maximums and proximity decay functions.
author:
 - name: Pukar Bhandari
   email: pukar.bhandari@wfrc.utah.gov
   affiliation:
     - name: Wasatch Front Regional Council
       url: "https://wfrc.utah.gov/"
date: "2025-12-15"
---

# Environment Setup

## Libraries

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak(
  c(
    "tidyverse",   # https://github.com/tidyverse/tidyverse/
    "sf",          # https://github.com/r-spatial/sf/
    "openrouteservice", # https://github.com/GIScience/openrouteservice-r/
    "DBI",         # https://github.com/r-dbi/DBI/
    "duckdb",      # https://github.com/duckdb/duckdb-r/
    "duckplyr",    # https://github.com/tidyverse/duckplyr/
    "duckspatial", # https://github.com/Cidree/duckspatial/
    "arrow",       # https://github.com/apache/arrow/tree/main/r
    "geoarrow",    # https://github.com/geoarrow/geoarrow-r/
    "janitor",     # https://github.com/sfirke/janitor/
    "mapgl",       # https://github.com/walkerke/mapgl/
    "here",        # https://github.com/r-lib/here/
    "fs",          # https://github.com/r-lib/fs/
    "zip"          # https://github.com/r-lib/zip
    ),
    dependencies = TRUE
  )
```

OR,

```r
install.packages("pak") # https://github.com/r-lib/pak

pak::pak("renv", dependencies = TRUE) # https://github.com/rstudio/renv/

renv::restore()
```

```{r}
#| label: setup-libraries
#| message: false
#| warning: false

library(dplyr)
library(DBI)
library(duckdb)
library(arrow)
library(geoarrow)
library(glue)
library(here)
library(fs)
```

# Configuration

```{r}
#| label: setup-dirs

dirs <- list(
  results   = here::here("_output"),
  processed = here::here("_data/processed"),
  base      = here::here("_data/processed", "base_layers"),
  parcels   = here::here("_data/processed", "parcels_taz")
)

# 1. INPUTS (Enriched Points from Step 3)
fp_pts_parcels_enriched <- here::here(dirs$results, "parcels_pt_enriched.parquet")
fp_pts_h3_enriched      <- here::here(dirs$results, "h3_pt_enriched.parquet")

# 2. GEOMETRY SOURCES (Polygons from Step 2)
fp_poly_parcels         <- here::here(dirs$parcels, "REMM_BY_parcels.parquet")
fp_poly_h3              <- here::here(dirs$base,    "Analysis_Boundary_H3_Grid.parquet")

# 3. OUTPUTS (Final Polygons)
fp_parcels_enriched <- here::here(dirs$results, "parcels_enriched.parquet")
fp_h3_enriched      <- here::here(dirs$results, "h3_enriched.parquet")

fp_parcels_scored   <- here::here(dirs$results, "parcels_scored.parquet")
fp_h3_scored        <- here::here(dirs$results, "h3_scored.parquet")
```

### Projection Constants

```{r}
#| label: setup-crs

# Projection Constants
CRS_STP <- "EPSG:3566"  # NAD83 / Utah Central (Projected - ftUS)
CRS_UTM <- "EPSG:26912" # NAD83 / UTM zone 12N (Projected - meters)
CRS_WGS <- "EPSG:4326"  # WGS84 (Geographic - Lat/Lon)
```

## Initialize Database

```{r}
#| label: init-db
conn <- DBI::dbConnect(duckdb::duckdb())

DBI::dbExecute(conn, "INSTALL spatial; LOAD spatial;")
DBI::dbExecute(conn, "CALL register_geoarrow_extensions()")

# Set safe memory limits
DBI::dbExecute(conn, "SET memory_limit = '40GB'; SET threads = 16;")
```

# Scoring & Weighting

## Scoring Logic

We define helper functions to standardize the scoring logic across both Parcels and H3 grids.

1. Identifies "static" columns (IDs, Admin codes) vs "scorable" columns.
2. Calculates the global MAX or MIN for normalization.
3. Applies **Inverse Normalization** ($\frac{Min}{Value}$) for Cost metrics (Time/Distance).
4. Applies **Linear Normalization** ($\frac{Value}{Max}$) for Benefit metrics (Jobs/Binary).
5. Exports the final table, replacing raw values with scores.

This function encapsulates the entire scoring workflow so it can be applied identically to both Parcels and H3 grids.

```{r}
#| label: func-scoring

score_points <- function(conn, input_path, table_name) {

  message(glue::glue("📊 Scoring {table_name}..."))

  # 1. Load Enriched Points
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE VIEW raw_{table_name} AS SELECT * FROM read_parquet('{input_path}')
  "))

  # 2. Identify Columns
  #    - Added "oid" so it is excluded from scoring.
  #    - Added "AT", "AA" to be safe.
  static_cols <- c("oid", "rowid", "geometry", "geom", "parcel_id", "h3_index", "CommCode", "COUNTYNBR", "BC")

  all_cols    <- DBI::dbListFields(conn, glue::glue("raw_{table_name}"))
  score_cols  <- setdiff(all_cols, static_cols)

  if(length(score_cols) == 0) stop("❌ No scorable columns found!")

  # 3. Calculate Scores (Normalization)
  sql_parts <- list()
  message("   ⚖️ Calculating Normalization Stats...")

  for (col in score_cols) {
    # Cost vs Benefit Detection
    is_cost <- grepl("^(TT_|AC|AH|AG|AM|AE_|TA_|TF)", col)

    if (is_cost) {
      # COST: Min / Value (Decay)
      val <- DBI::dbGetQuery(conn, glue::glue('SELECT MIN("{col}") as m FROM raw_{table_name} WHERE "{col}" > 0'))$m
      if (is.na(val) || val == 0) val <- 1.0
      sql_parts[[col]] <- glue::glue('CASE WHEN "{col}" IS NULL OR "{col}" = 0 THEN 0.0 WHEN "{col}" <= {val} THEN 1.0 ELSE {val} / "{col}" END AS "{col}"')
    } else {
      # BENEFIT: Value / Max (Linear)
      val <- DBI::dbGetQuery(conn, glue::glue('SELECT MAX("{col}") as m FROM raw_{table_name}'))$m
      if (is.na(val) || val == 0) val <- 1.0
      sql_parts[[col]] <- glue::glue('CASE WHEN "{col}" IS NULL THEN 0.0 ELSE "{col}" / {val} END AS "{col}"')
    }
  }

  # 4. Composite Logic (Weighted Groups)
  composites <- list()
  if (all(c("TT_Local", "TT_Rapid", "TT_CRT") %in% score_cols)) {
    composites[["TT"]] <- "GREATEST(COALESCE(TT_Local,0)*0.5, COALESCE(TT_Rapid,0)*0.8, COALESCE(TT_CRT,0)*1.0) AS TT"
  }
  if (all(c("AE_K12", "AE_High") %in% score_cols)) {
    composites[["AE"]] <- "GREATEST(COALESCE(AE_K12,0)*0.8, COALESCE(AE_High,0)*1.0) AS AE"
  }
  if (all(c("TA_class1", "TA_class2", "TA_trails") %in% score_cols)) {
     composites[["TA"]] <- "GREATEST(COALESCE(TA_class2,0)*0.5, COALESCE(TA_class1,0)*1.0, COALESCE(TA_trails,0)*1.0) AS TA"
  }

  # 5. Build View (WITH QUOTED COLUMNS)
  #    We explicitly wrap static columns in double quotes to handle "AT", "AA", etc.
  safe_static_cols <- paste0('"', intersect(all_cols, static_cols), '"')
  select_static    <- paste(safe_static_cols, collapse = ", ")

  select_scores    <- paste(unlist(sql_parts), collapse = ", \n      ")

  # Handle composites and comma safely
  if (length(composites) > 0) {
    select_comps <- paste(",", paste(unlist(composites), collapse = ", \n      "))
  } else {
    select_comps <- ""
  }

  # Create SCORED View
  # Note: select_comps includes the leading comma if composites exist.
  query <- glue::glue("
    CREATE OR REPLACE TABLE scored_{table_name} AS
    WITH base_norm AS (
      SELECT {select_static}, {select_scores}
      FROM raw_{table_name}
    )
    SELECT * {select_comps} FROM base_norm
  ")

  DBI::dbExecute(conn, query)
  DBI::dbExecute(conn, glue::glue("DROP VIEW raw_{table_name}"))
  message("   ✅ Scores Calculated (In-Memory)")
}
```

## Run Scoring

```{r}
#| label: run-parcels

# Calculate scores for both layers (Result is stored in DuckDB tables 'scored_parcels' and 'scored_h3')
score_points(conn, fp_pts_parcels_enriched, "parcels")
score_points(conn, fp_pts_h3_enriched, "h3")
```

## Rejoin Geometry & Export

This step joins the lightweight scored data back to the heavy Polygons and saves the final result.

```{r}
#| label: func-export

join_and_export <- function(conn, geom_path, point_table, id_col, out_path) {

  message(glue::glue("🔗 Joining {point_table} to Polygons -> {basename(out_path)}..."))

  # 1. Load Polygon Geometry & Standardize
  #    Using the 'Universal Adapter' (ST_GeomFromWKB + ST_AsWKB) ensures this works
  #    regardless of whether the input is GeoParquet (Geometry) or Raw Parquet (Blob).
  DBI::dbExecute(conn, glue::glue("
    CREATE OR REPLACE TEMP TABLE raw_polys AS
    SELECT
      {id_col},
      ST_GeomFromWKB(ST_AsWKB(geometry)) as poly_geom
    FROM read_parquet('{geom_path}')
  "))

  # 2. Join & Export
  #    - We select ALL attributes from the point table (pts.*)
  #    - We EXCLUDE the point geometry column ('geom') so it doesn't conflict
  #    - We join and transform the polygon geometry ('poly_geom') from the boundary file
  DBI::dbExecute(conn, glue::glue("
    COPY (
      SELECT
        pts.* EXCLUDE (geom, oid), -- Drop the Point geometry

        -- Transform Polygon to WGS84 and export as WKB 'geometry' column
        ST_AsWKB(ST_Transform(
           poly.poly_geom,
           '{CRS_UTM}', '{CRS_WGS}', always_xy := true
        )) as geometry

      FROM {point_table} pts
      JOIN raw_polys poly ON pts.{id_col} = poly.{id_col}
    ) TO '{out_path}' (FORMAT PARQUET)
  "))

  DBI::dbExecute(conn, "DROP TABLE raw_polys")
}
```

```{r}
#| label: export-scored

# Join Scored Points -> Polygons
join_and_export(conn, fp_poly_parcels, "scored_parcels", "parcel_id", fp_parcels_scored)
join_and_export(conn, fp_poly_h3,      "scored_h3",      "h3_index",  fp_h3_scored)
```

```{r}
#| label: export-enriched
# We also want the RAW values (minutes/miles) attached to polygons for debugging/analysis.
# So we reload the 'enriched' points and join them too.

message("🔗 Joining Raw Enriched Data to Polygons...")

# Parcels
DBI::dbExecute(conn, glue::glue("CREATE OR REPLACE VIEW raw_pts_parcels AS SELECT * FROM read_parquet('{fp_pts_parcels_enriched}')"))
join_and_export(conn, fp_poly_parcels, "raw_pts_parcels", "parcel_id", fp_parcels_enriched)

# H3
DBI::dbExecute(conn, glue::glue("CREATE OR REPLACE VIEW raw_pts_h3 AS SELECT * FROM read_parquet('{fp_pts_h3_enriched}')"))
join_and_export(conn, fp_poly_h3, "raw_pts_h3", "h3_index", fp_h3_enriched)
```

### Export to GDB

```{r}
#| label: export-gdb

message("🌍 Converting Parquets to File Geodatabase...")

# 1. Configuration
gdb_path <- here::here(dirs$results, "HousingSuitability_Centers202512.gdb")

# Define inputs: Layer Name -> File Path
to_export <- list(
  "Parcels_Enriched" = fp_parcels_enriched,
  "H3_Enriched"      = fp_h3_enriched,
  "Parcels_Scored"   = fp_parcels_scored,
  "H3_Scored"        = fp_h3_scored
)

# 2. Execution
purrr::iwalk(to_export, function(path, layer_name) {

  if (!fs::file_exists(path)) return(message(glue::glue("⚠️ Skipping {layer_name} (Not found)")))

  message(glue::glue("  ↳ Processing {layer_name}..."))

  # Streamlined Approach: Open -> Convert to SF -> Write
  arrow::open_dataset(path) |>
    sf::st_as_sf(crs = CRS_WGS) |> # WGS84
    sf::st_cast("MULTIPOLYGON") |>
    # Optional: Uncomment next line if you need the GDB in UTM
    # sf::st_transform(CRS_UTM) |>
    sf::st_write(dsn = gdb_path, layer = layer_name, driver = "OpenFileGDB",
                 append = FALSE, quiet = TRUE)
})

message("✅ GDB Export Complete!")
```

```{r}
#| label: zip-gdb

message("📦 Compressing GDB to Zip...")

# 3. Create Zip
zip::zip(
  zipfile = paste0(gdb_path, ".zip"),  # Where to save the result (Absolute path works best)
  files   = fs::path_file(gdb_path), # What to grab (Relative to 'root')
  root    = fs::path_dir(gdb_path),  # The virtual working directory for this operation
  mode    = "mirror"       # "mirror" is best for exact copies
)

message(glue::glue("✅ Compressed successfully."))
```

## Validation

Check a few records to ensure scores make sense (e.g., scores shouldn't be > 1.0 or infinite).

```{r}
#| label: validate
#| echo: false

arrow::open_dataset(fp_parcels_scored) |> head() |> collect()
```

## Cleanup

```{r}
#| label: cleanup
DBI::dbDisconnect(conn, shutdown = TRUE)
```
